---
title: "FL100 FFQ Lactose and DTOTAL"
---

Author: E. Chin
Date: 02/07/2022
Purpose: Get Lactose and D_TOTAL from FFQ for use for 16S <> LCT <> Lactose paper 



```{r}
library(dplyr)
setwd("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/")
ffq<- read.csv("raw_ffq/CTSC24532USDAWHNRCNu_DATA_2022-02-07_1003.csv", header=T)
lct<- read.csv("/Users/elizabeth.chin/Desktop/milk/genotype_ethnicity/FL100_FINAL_LCT_genotype_clean.csv", header=T)
eth<- read.csv("/Users/elizabeth.chin/Desktop/milk/genotype_ethnicity/FL100_FINAL_ethnicities_clean.csv", header=T)
bins<-read.csv("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/CTSC24532USDAWHNRCNu-BinAgeSexBMIallSubje_DATA_2022-02-09_1449.csv", header=T)

#bins file came from Ellen Bonnel emailed 02/09/2022; age is v1 (from physiology), sex from physiology, bmi_final is calculated using visit 1 height and visit 2 (fasting) weight

```

```{r}
dim(ffq) #393 x 1081
#find cols related to qc
colnames(ffq)[1000:1081]
```

get just the dairy columns, QC columns   

lacs is in grams   
dtotal in cup eq. from FFQ data dict: Total milk, yogurt, cheese, and whey. For some foods, the total dairy values
could be higher than the sum of D_MILK, D_YOGURT, and D_CHEESE
because Miscellaneous dairy component composed of whey which is not
included in FPED as a separate variable.   


```{r}
dairy<- ffq %>% select(subject_id, d_total, lacs, dt_kcal, ffq_qaqc_kcalok:block_ffq_complete)
```

ffq_qc_passed is the variable created for whether a subject passed QC after Joanne Arsenault did QC (pass = 1, 0 = no); includes consideration for total intake and number of foods reported 
```{r}
sum(is.na(dairy)) #168
plyr::count(dairy, "ffq_qc_passed")
```

```{r}
passing <- dairy %>% filter(ffq_qc_passed == 1) %>% select(-c(ffq_qaqc_kcalok:block_ffq_complete)) #359 subjects, correct 
```

merge with the metadata
```{r}
master<- merge(merge(lct, eth, by = "SubjectID"), bins, by.x = "SubjectID", by.y="subject_id") %>%
  dplyr::select(-c(visit1_date:age_ffq)) #349 subjects

```
```{r}
df<- merge(master, passing, by.x = "SubjectID", by.y= "subject_id") #346 subjects
```




```{r}
setwd("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/")

#save all dtotal and lacs for those passing ffq
write.csv(passing, "out_ffq/01_ffq_lactose_dtotal.csv", row.names=F)
write.csv(df, "out_ffq/02_ffq_lactose_dtotal_lct.csv", row.names=F)
```

