---
title: "01_ASA24_LNP_analysis"
---

Author: E.Chin
Purpose: ASA24 Lactose and D_TOTAL analysis w.r.t. LNP genotypes; for use with 16s analysis paper. 

Note 02/11/2022- we will exclude LNP african americans from all-ethnicity analysis since we cannot be sure of their LP status. Multi-ethnic African American/Black participants are 6083 and 8015 

```{r setup}
library(dplyr)
library(ggplot2)
library(car)
library(doBy)
library(GGally)
library(corrplot)
library(cowplot)
library(emmeans)

asa<- read.csv("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/out_asa24/03_asa24_lactose_dtotal_no_powders_lct.csv", header=T) #these are the subjects who have lactose (average over 2 or 3 recalls) estimates available from usable ASA24 recalls  

#remove African American subjects who are GG
af_gg<- asa %>% filter(Ethnicity == "African.American" & LCT == "GG")
'%nin%'<-Negate('%in%')
asa_filt<- asa %>% filter(SubjectID %nin% af_gg$SubjectID) %>% filter(SubjectID != 6083 & SubjectID != 8015)
plyr::count(asa_filt, c("Ethnicity", "LCT"))
asa <- asa_filt
```

```{r}
summary(asa)
```

```{r}
asa$sex<- as.factor(asa$sex)
asa$Ethnicity<-as.factor(asa$Ethnicity)
asa$LCT<-as.factor(asa$LCT)
asa$LNP<-as.factor(asa$LNP)
```

Does the frequency of LP vs LNP significantly differ among ethnicities? 

```{r}
cnts<- plyr::count(asa, c("Ethnicity", "LNP"))
cnts_wide <- tidyr::spread(cnts, LNP, freq)
cnts_wide[is.na(cnts_wide)]<- 0
rownames(cnts_wide)<- cnts_wide$Ethnicity
cnts_wide<- cnts_wide[-1,-1] #get rid of Ethnicity and African Americans since that is artificailly made to have LNP=0
fisher.test(as.matrix(cnts_wide)) # 2.2e-16

pairwiseNominalIndependence(as.matrix(cnts_wide), fisher = T, gtest=F, chisq=F, method = "fdr") %>% htmlTable::htmlTable(rnames = F)

```

```{r}
cnts_wide %>% htmlTable::htmlTable()
```


correlation matrix
```{r}
asatmp<- asa %>% dplyr::select(SubjectID, age, ht_cm:bmi_final, Lactose.consumed:KCAL)
asacorr<- cor(asatmp[,-1])
asap<- ggcorrplot::cor_pmat(asatmp[,-1])
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/corrplot_asa.pdf", height = 5, width = 6)
corrplot::corrplot(asacorr, is.corr=F, tl.srt=65, p.mat=asap, sig.level = 0.05, insig="pch",
                   mar=c(0,0,0,0))
dev.off()
```
```{r}
asacorr
asap
```


plots
```{r, eval =F}

ggduo(asa, 
      mapping=aes(color=LNP),
      columnsX = c("Lactose.consumed","D_TOTAL", "KCAL"),
      columnsY=c("LNP", "Ethnicity", "age", "sex"),
      types = list(continuous= "points", comboHorizontal="box"))

ggduo(asa, 
      mapping=aes(color=LNP),
      columnsX = c("Lactose.consumed","D_TOTAL", "KCAL"),
      columnsY=c("ht_cm", "weightv2", "bmi_final"),
      types = list(continuous= "points", comboHorizontal="box")) +
  theme_bw()

```
```{r}
a<- ggplot(data = asa, aes(x = KCAL, y = Lactose.consumed, color = LNP)) + geom_point(aes(shape = sex), size = 2) + 
  geom_rug() + theme_bw() + theme(legend.position = "bottom") + ggtitle("Latose (g) vs KCAL")
b<- ggplot(data = asa, aes(x = KCAL, y = D_TOTAL, color = LNP)) + geom_point(aes(shape = sex), size = 2) + geom_rug() +theme_bw() + theme(legend.position = "bottom") + ggtitle("Total Dairy (cup eq.) vs KCAL")
#pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/rugplot_asa.pdf", width = 7, height = 5)
plot_grid(a, b)
dev.off()
```
# Check out distributions

```{r}
hist(asa$Lactose.consumed)
hist(asa$D_TOTAL)
hist(log(asa$D_TOTAL))
```


transform lactose
```{r transform-lactose}
library(MASS)
Box = boxcox(asa$Lactose.consumed ~ 1,              # Transform Turbidity as a single vector
             lambda = seq(-6,6,0.1))      # Try values -6 to 6 by 0.1
Cox = data.frame(Box$x, Box$y)            # Create a data frame with the results
Cox2 = Cox[with(Cox, order(-Cox$Box.y)),] # Order the new data frame by decreasing y
Cox2[1,]                                  # Display the lambda with the greatest
                                          #    log likelihood
lambda = Cox2[1, "Box.x"]                 # Extract that lambda 0.3
lac_box = (asa$Lactose.consumed ^ lambda - 1)/lambda   # Transform the original data
hist(lac_box)
shapiro.test(lac_box) #p-value = 0.06158
asa$lac_box<-lac_box
```

transform d_total
```{r transform-dtotal}
#there are some 0s 
library(rcompanion)
total_tukey<- transformTukey(asa$D_TOTAL)
asa$total_tukey<-total_tukey

```
transform kcal

NOTE- we will use a lambda= 0.3 to transform KCAL to keep the lambdas the same between kcal and lactose; this will make back transformation of emmeans easier. 

```{r}
asa$kcal_box<- (asa$KCAL ^ 0.3 - 1) / 0.3
```

```{r, eval = F}
Box = boxcox(asa$KCAL ~ 1,    # Transform as a single vector
             lambda = seq(-6,6,0.1))      # Try values -6 to 6 by 0.1
Cox = data.frame(Box$x, Box$y)            # Create a data frame with the results
Cox2 = Cox[with(Cox, order(-Cox$Box.y)),] # Order the new data frame by decreasing y
Cox2[1,]                                  # Display the lambda with the greatest
                                          #    log likelihood
lambda = Cox2[1, "Box.x"]                 # Extract that lambda 0.4
kcal_box = (asa$KCAL ^ lambda - 1)/lambda   # Transform the original data
hist(kcal_box)
shapiro.test(kcal_box) #p-value = .6298
asa$kcal_box<-kcal_box
```
# Are there differences among sex and ethnicity? 

```{r}
make_boxplots<- function(x, y, dat, y_transformed){
  varx = dat[,x]
  vary = dat[,y]
  y_trans = dat[,y_transformed]
  p<- ggplot(data = dat, aes(x = varx, y = vary)) + geom_boxplot(aes(fill = varx), varwidth = T, notch = T) + theme_bw() +
    xlab(x) + ylab(y)
  print(p)
  fit<-lm(y_trans~varx, data = dat)
  plot(fit)
  #summary(fit) %>% print()
  car::Anova(fit) %>% print()
}

##total intake does not significantly differ by ethnicities
make_boxplots(dat = asa, y = "KCAL", x = "Ethnicity", y_transformed = "kcal_box")
make_boxplots(dat = asa, y = "D_TOTAL", x = "Ethnicity", y_transformed = "total_tukey") #0.008738
make_boxplots(dat = asa, y = "Lactose.consumed", x = "Ethnicity", y_transformed = "lac_box")#0.008545
##since ethnicity has super unequal groups let's do a sanity check with kruskal
kruskal.test(asa$KCAL ~ asa$Ethnicity) #0.2806
kruskal.test(asa$D_TOTAL ~ asa$Ethnicity) #0.008461
kruskal.test(asa$Lactose.consumed ~asa$Ethnicity) #0.008461

make_boxplots(dat = asa, y = "KCAL", x = "sex", y_transformed = "kcal_box") 
#men consumed significantly more kcal than women; p-value: < 2.2e-16
make_boxplots(dat = asa, y = "D_TOTAL", x = "sex", y_transformed = "total_tukey") 
#men consumed significantly more total dairy than women; p-value = 0.003952
make_boxplots(dat = asa, y = "Lactose.consumed", x = "sex", y_transformed = "lac_box")
#difference in consumption approached significance; p-value 0.06978
```
```{r}
a<- ggplot(data = asa, aes(x = Ethnicity, y = Lactose.consumed)) + geom_boxplot(aes(fill=Ethnicity)) + theme_bw() + theme(legend.position = "none", axis.text.x = element_text(angle = 45, vjust=1, hjust = 1))
b<- ggplot(data = asa, aes(x = Ethnicity, y = D_TOTAL)) + geom_boxplot(aes(fill=Ethnicity)) + theme_bw() + theme(legend.position  = "none", axis.text.x = element_text(angle = 45, vjust=1, hjust = 1))
#pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/boxplot_asa_lactose_dtotal_eth.pdf", width = 7, height = 5)
plot_grid(a,b)
dev.off()
```


```{r plot-lactosevdtotal}
p<-ggplot(data=asa, aes(x= D_TOTAL, y = Lactose.consumed)) + geom_point(aes(shape=sex, color = LNP), size = 2) + theme_bw() + 
  ggtitle("Average Lactose Consumption vs Dairy Intake ") + ylab("Estimated Lactose Consumption (g)") +
  xlab("Total Dairy Consumption (servings)") + geom_sasath(method="lm", se=F, col= "darkgray", linetype="dashed") + 
  theme(legend.position = "bottom", 
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank())
p
#ggsave(p,file = "/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/scatter_asa_lactose_dtotal.pdf", height=5, width=6, units="in", dpi=300)
```

# Does Lactose consumption significantly differ between LP and LNP (all subjects)? 

```{r}
lac_fit<- lm(lac_box ~ kcal_box+Ethnicity+LNP, data = asa) #we can't be sure about any LNP*Ethnicity interaction because we have taken out all GG African Americans, so we keep everything as + 
plot(lac_fit) #looks good
hist(lac_fit$residuals) #looks good
summary(lac_fit)
Anova(lac_fit)
```

```{r eval = F}
#keep this as a note, in case you ever want to do back calculation on the emmeans 
bctran <- make.tran("boxcox", 0.3)
lac_bc <- with(bctran, 
                lm(linkfun(Lactose.consumed) ~ linkfun(KCAL) + LNP, data = asa))
plot(lac_bc)
hist(lac_bc$residuals)
emmeans(lac_bc, "LNP")
emmeans(lac_bc, "LNP", type ="response")
```


Does total dairy consumption significantly differ between LP and LNP subjects?

```{r}
#simple t test
t.test(lac_box ~ LNP, data = asa) #0.001223
t.test(total_tukey ~ LNP, data = asa)#0.009083

```

```{r}
dairy_fit <- lm(total_tukey ~ kcal_box + Ethnicity +LNP, data = asa)
plot(dairy_fit)
hist(dairy_fit$residuals)
summary(dairy_fit)
Anova(dairy_fit)
```

```{r}
plyr::count(asa, c("Ethnicity", "LNP")) %>% htmlTable::htmlTable()
```

```{r eval = F}
a<- ggplot(data = asa, aes(x = LNP, y = Lactose.consumed)) + geom_boxplot(aes(fill = LNP), varwidth=T, notch=T ) + theme_bw() + theme(legend.position ="none")
b<- ggplot(data = asa, aes(x = LNP, y = D_TOTAL)) + geom_boxplot(aes(fill = LNP), varwidth=T, notch=T ) + theme_bw() + theme(legend.position ="none")
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/boxplots_asa_lac_dairy_lnp.pdf", width = 7, height = 5)
plot_grid(a, b)
dev.off()
```
Average consumption (raw data, not the estimated/adjusted mean)
```{r}
asa %>% group_by(Ethnicity, LNP) %>% 
  summarize(mean_lactose = round(mean(Lactose.consumed),2),
            sd_lactose = round(sd(Lactose.consumed),2),
            mean_dairy = round(mean(D_TOTAL),2),
            sd_dairy = round(sd(D_TOTAL),2)) %>% htmlTable::htmlTable(rnames = F)
```


