---
title: "02_ASA24_LNP_analysis_caucasians_hispanics"
---

Author: E.Chin
Purpose: ASA24 Lactose and D_TOTAL analysis w.r.t. LNP genotypes; for use with 16s analysis paper. This is for only the Caucasian and Hispanic subjects. 


```{r setup}
library(dplyr)
library(ggplot2)
library(car)
library(doBy)
library(GGally)
library(corrplot)
library(cowplot)
library(emmeans)

asa<- read.csv("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/out_asa24/03_asa24_lactose_dtotal_no_powders_lct.csv", header=T) #these are the subjects who have lactose (average over 2 or 3 recalls) estimates available from usable ASA24 recalls  

#Only keep the Caucasian and Hispanic subjects
asa<-asa %>% filter(Ethnicity == "Caucasian" | Ethnicity == "Hispanic")
```

```{r}
summary(asa)
dim(asa)
```

```{r}
asa$sex<- as.factor(asa$sex)
asa$Ethnicity<-as.factor(asa$Ethnicity)
asa$LCT<-as.factor(asa$LCT)
asa$LNP<-as.factor(asa$LNP)
```

correlation matrix
```{r}
asatmp<- asa %>% dplyr::select(SubjectID, age, ht_cm:bmi_final, Lactose.consumed:KCAL)
asacorr<- cor(asatmp[,-1])
asap<- ggcorrplot::cor_pmat(asatmp[,-1])
#pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/corrplot_asa_CH.pdf", height = 5, width = 6)
corrplot::corrplot(asacorr, is.corr=F, tl.srt=65, p.mat=asap, sig.level = 0.05, insig="pch",
                   mar=c(0,0,0,0))
dev.off()
```
```{r}
asacorr
asap
```


Summary of intake by genotype
```{r}
aa<- asa %>% filter(LCT == "AA") %>% summary(.)
ag<- asa %>% filter(LCT == "AG") %>% summary(.)
gg<- asa %>% filter(LCT == "GG") %>% summary(.)

summaryBy(Lactose.consumed ~ LCT, data=asa, FUN = list(mean, max, min, median)) %>%
  htmlTable::htmlTable()
```


plots
```{r, eval =F}

ggduo(asa, 
      mapping=aes(color=LNP),
      columnsX = c("Lactose.consumed","D_TOTAL", "KCAL"),
      columnsY=c("LNP", "Ethnicity", "age", "sex"),
      types = list(continuous= "points", comboHorizontal="box"))

ggduo(asa, 
      mapping=aes(color=LNP),
      columnsX = c("Lactose.consumed","D_TOTAL", "KCAL"),
      columnsY=c("ht_cm", "weightv2", "bmi_final"),
      types = list(continuous= "points", comboHorizontal="box")) +
  theme_bw()

```
```{r}
a<- ggplot(data = asa, aes(x = KCAL, y = Lactose.consumed, color = LNP)) + geom_point(aes(shape = sex), size = 2) + 
  geom_rug() + theme_bw() + theme(legend.position = "bottom") + ggtitle("Latose (g) vs KCAL")
b<- ggplot(data = asa, aes(x = KCAL, y = D_TOTAL, color = LNP)) + geom_point(aes(shape = sex), size = 2) + geom_rug() +theme_bw() + theme(legend.position = "bottom") + ggtitle("Total Dairy (cup eq.) vs KCAL")
#pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/rugplot_asa.pdf", width = 7, height = 5)
plot_grid(a, b)
dev.off()
```
# Check out distributions

```{r}
hist(asa$Lactose.consumed)
hist(asa$D_TOTAL)
hist(log(asa$D_TOTAL))
```


transform lactose
```{r transform-lactose}
library(MASS)
Box = boxcox(asa$Lactose.consumed ~ 1,              # Transform Turbidity as a single vector
             lambda = seq(-6,6,0.1))      # Try values -6 to 6 by 0.1
Cox = data.frame(Box$x, Box$y)            # Create a data frame with the results
Cox2 = Cox[with(Cox, order(-Cox$Box.y)),] # Order the new data frame by decreasing y
Cox2[1,]                                  # Display the lambda with the greatest
                                          #    log likelihood
lambda = Cox2[1, "Box.x"]                 # Extract that lambda 0.3
lac_box = (asa$Lactose.consumed ^ lambda - 1)/lambda   # Transform the original data
hist(lac_box)
shapiro.test(lac_box) #p-value = 0.06158
asa$lac_box<-lac_box
```

transform d_total
```{r transform-dtotal}
#there are some 0s 
library(rcompanion)
total_tukey<- transformTukey(asa$D_TOTAL)
asa$total_tukey<-total_tukey

```
transform kcal

NOTE- we will use a lambda= 0.3 to transform KCAL to keep the lambdas the same between kcal and lactose; this will make back transformation of emmeans easier. 
the 'best' lambda by searching below is 0.2

```{r}
asa$kcal_box<- (asa$KCAL ^ 0.3 - 1) / 0.3
```

```{r, eval = F}
Box = boxcox(asa$KCAL ~ 1,    # Transform as a single vector
             lambda = seq(-6,6,0.1))      # Try values -6 to 6 by 0.1
Cox = data.frame(Box$x, Box$y)            # Create a data frame with the results
Cox2 = Cox[with(Cox, order(-Cox$Box.y)),] # Order the new data frame by decreasing y
Cox2[1,]                                  # Display the lambda with the greatest
                                          #    log likelihood
lambda = Cox2[1, "Box.x"]                 # Extract that lambda 0.2
kcal_box = (asa$KCAL ^ lambda - 1)/lambda   # Transform the original data
hist(kcal_box)
shapiro.test(kcal_box) #p-value = .6298
asa$kcal_box<-kcal_box
```
# Are there differences among sex and ethnicity? 

```{r}
make_boxplots<- function(x, y, dat, y_transformed){
  varx = dat[,x]
  vary = dat[,y]
  y_trans = dat[,y_transformed]
  p<- ggplot(data = dat, aes(x = varx, y = vary)) + geom_boxplot(aes(fill = varx), varwidth = T, notch = T) + theme_bw() +
    xlab(x) + ylab(y)
  print(p)
  fit<-lm(y_trans~varx, data = dat)
  plot(fit)
  #summary(fit) %>% print()
  car::Anova(fit) %>% print()
}

#no significant differences among ethnicities
make_boxplots(dat = asa, y = "KCAL", x = "Ethnicity", y_transformed = "kcal_box")#0.9163
make_boxplots(dat = asa, y = "D_TOTAL", x = "Ethnicity", y_transformed = "total_tukey") #0.5005
make_boxplots(dat = asa, y = "Lactose.consumed", x = "Ethnicity", y_transformed = "lac_box")#0.258


make_boxplots(dat = asa, y = "KCAL", x = "sex", y_transformed = "kcal_box") 
#men consumed significantly more kcal than women; p-value: 1.993e-15
make_boxplots(dat = asa, y = "D_TOTAL", x = "sex", y_transformed = "total_tukey") 
#men consumed significantly more total dairy than women; p-value = 0.001864 
make_boxplots(dat = asa, y = "Lactose.consumed", x = "sex", y_transformed = "lac_box")
#difference in consumption approached significance; p-value 0.06473
```
```{r}
a<- ggplot(data = asa, aes(x = Ethnicity, y = Lactose.consumed)) + geom_boxplot(aes(fill=Ethnicity)) + theme_bw() + theme(legend.position = "none", axis.text.x = element_text(angle = 45, vjust=1, hjust = 1))
b<- ggplot(data = asa, aes(x = Ethnicity, y = D_TOTAL)) + geom_boxplot(aes(fill=Ethnicity)) + theme_bw() + theme(legend.position  = "none", axis.text.x = element_text(angle = 45, vjust=1, hjust = 1))
plot_grid(a,b)
```


```{r plot-lactosevdtotal}
p<-ggplot(data=asa, aes(x= D_TOTAL, y = Lactose.consumed)) + geom_point(aes(shape=sex, color = LNP), size = 2) + theme_bw() + 
  ggtitle("Average Lactose Consumption vs Dairy Intake ") + ylab("Estimated Lactose Consumption (g)") +
  xlab("Total Dairy Consumption (servings)") + geom_smooth(method="lm", se=F, col= "darkgray", linetype="dashed") + 
  theme(legend.position = "bottom", 
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank())
p
#ggsave(p,file = "/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/scatter_asa_lactose_dtotal.pdf", height=5, width=6, units="in", dpi=300)
```

# Does Lactose consumption significantly differ between LP and LNP (all subjects)? 
```{r}
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/interaction_asa_lac_kcal_CH.pdf", width = 4, height =3)
ggplot(data = asa, aes(x = KCAL, y = Lactose.consumed, color = LNP)) + geom_point(aes(fill = LNP)) + 
  geom_smooth(method = "lm", se = F, aes(color = LNP)) + theme_bw()
dev.off()
ggplot(data = asa, aes(x = KCAL, y = D_TOTAL, color = LNP)) + geom_point(aes(fill = LNP)) + 
  geom_smooth(method = "lm", se = F, aes(color = LNP)) + theme_bw()
```

adapted from https://stats.oarc.ucla.edu/r/seminars/interactions-r/#s4 
```{r}
bctran <- make.tran("boxcox", 0.3)
lac_bc <- with(bctran, 
                lm(linkfun(Lactose.consumed) ~ linkfun(KCAL)*LNP, data = asa))
summary(lac_bc)



lac_fit<- lm(lac_box ~ kcal_box * LNP, data = asa)
plot(lac_fit) #looks good
hist(lac_fit$residuals) #looks good
summary(lac_fit) #LNP is the reference level
# intercept: when kcal_box is 0 and the ref group is LNP then lac_box is 1.62
# b1 kcal-box: the simple slope for kcal_box for LNP is 0.037 
# b2 LP: the simple effect of LNP status (the diference in lac_box between LP and LNP when kcal = 0) is -4.52
# interaction: the difference in the simple slopes of kcal_box for LP vs LNP 
```
#getting the simple slopes for each level of the categorical moderator (LNP)
```{r}
emtrends(lac_bc, ~LNP, var = "KCAL")
#the simple slope for LNP is 0.037 (same as b1) and for males it is 0.206 = 0.037 + 0.169; the 95% CI contains 0 for LNP so the simple slope is significant for LNP but not LP
emtrends(lac_fit, pairwise ~ LNP, var="kcal_box")
emmip(lac_fit, LNP ~ kcal_box, cov.reduce = range)

```

```{r}
emtrends(lac_bc, pairwise~LNP, var = "KCAL", type = "response")
emmip(lac_bc, LNP ~ KCAL, cov.reduce = range)

```


Does total dairy consumption significantly differ between LP and LNP subjects?
```{r}
dairy_fit <- lm(total_tukey ~ kcal_box+LNP, data = asa) #no significant interaction, so we keep the simple model
plot(dairy_fit)
hist(dairy_fit$residuals)
summary(dairy_fit)
Anova(dairy_fit)
```
since it's kind of difficult to interpret from a biological perspective, whether lactose consumption was different between LP and LNP, we'll do a simple t test. in the discussion, it makes sense to talk about this first then get into the interactions

```{r}
t.test(data = asa, lac_box ~ LNP, alternative = "two.sided") #0.01378
t.test(data = asa, total_tukey ~ LNP, alternative = "two.sided") #0.08826
```


```{r eval = F}
a<- ggplot(data = asa, aes(x = LNP, y = Lactose.consumed)) + geom_boxplot(aes(fill = LNP), varwidth=T, notch=T ) + theme_bw() + theme(legend.position ="none")
b<- ggplot(data = asa, aes(x = LNP, y = D_TOTAL)) + geom_boxplot(aes(fill = LNP), varwidth=T, notch=T ) + theme_bw() + theme(legend.position ="none")
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/boxplots_asa_lac_dairy_lnp_CH.pdf", width = 7, height = 5)
plot_grid(a, b)
dev.off()
```
Average consumption (raw data, not the estimated/adjusted mean)
```{r}
asa %>% group_by(LNP) %>% 
  summarize(mean_lactose = round(mean(Lactose.consumed),2),
            mean_dairy = round(mean(D_TOTAL),2)) %>% htmlTable::htmlTable()
```

###

```{r}
fit<- lm(lac_box ~ LNP, data = asa)
plot(fit)
summary(fit)
stats::confint(fit)

fit_controlled <- lm(lac_box ~ LNP+kcal_box, asa) #no significant interaction so keep it as + only
plot(fit_controlled)
summary(fit_controlled)
stats::confint(fit_controlled)

```



does lactose intake differ between men and women?    
(expectation: yes, because men more likely to have overall greater intake than women)    

Lactose intake does not significantly differ between men and women (p = 0.07, two sided t test)
```{r lactose-sex}
t.test(lac_box ~ sex, data=asa) #p-value = 0.07436
ggplot(data=asa, aes(x= sex, y = Lactose.consumed)) + geom_boxplot(varwidth=T, notch=T, aes(fill=sex)) + theme_bw()

```



Does total dairy differ between men and women?
```{r dtotal-sex}
fit<-(lm(total_tukey ~ LNP*factor(sex), data=asa))
plot(fit)

Anova(lm(lac_box ~ sex*LCT, data = asa))

Anova(lm(lac_box ~ kcal_box + LNP, data = asa %>% filter(sex == 1)))
```



```{r scatterplot-metadata}
for(i in 5:9){
  #filepath=paste("/Users/elizabeth.chin/Desktop/milk/DXA/plots_102419/genotype_sex/","genotypesex_vs_",colnames(cd)[i],".png",sep="")
  k<- colnames(asa[i])
  d<-data.frame(Subject=factor(asa$SubjectID),
                Genotype=asa$LCT,
                var=asa[,i],
                Lactose=asa$Lactose.consumed,
                Sex=factor(asa$sex))
  q<- ggplot(d, aes(x= var, y= Lactose)) + 
    geom_point(aes(color=Genotype, shape = Sex), size=3)+
    geom_sasath(method="lm", se=F)+
    theme_bw() + xlab(print(k)) + 
    ylab("Lactose") +
    #facet_wrap(~Sex)+
    ggtitle(paste0(k," vs Lactose")) +
    #stat_summary(fun.y=mean, geom="point", shape=17, size=4)+
    theme(plot.title = element_text(size=22),
          axis.title.x = element_text(size=22), 
          axis.title.y = element_text(size=22))
  print(q) 
  #ggsave(file=filepath, plot=q, width=10, height=10)
}

for(i in c(5,7:9)){
  #filepath=paste("/Users/elizabeth.chin/Desktop/milk/DXA/plots_102419/genotype_sex/","genotypesex_vs_",colnames(cd)[i],".png",sep="")
  k<- colnames(asa[i])
  d<-data.frame(Subject=factor(asa$SubjectID),
                Genotype=asa$LCT,
                var=asa[,i],
                Lactose=asa$Lactose.consumed,
                Sex=factor(asa$sex))
  q<- ggplot(d, aes(x= var, y= Lactose)) + 
    geom_point(aes(color=Genotype), size=3)+
    geom_sasath(method= "lm", se=F) +
    theme_bw() + xlab(print(k)) + 
    ylab("Lactose") +
    facet_wrap(~Sex, scales="free_x")+
    ggtitle(paste0(k," vs Lactose")) +
    #stat_summary(fun.y=mean, geom="point", shape=17, size=4)+
    theme(plot.title = element_text(size=22),
          axis.title.x = element_text(size=22), 
          axis.title.y = element_text(size=22))
  print(q)
  #ggsave(file=filepath, plot=q, width=10, height=10)
}
```

```{r corrplot-lactose-meta-all}
library(magicfor)
magic_for()
fe<- asa %>% filter(sex == "female")
ma<- asa %>% filter(sex=="male")
for(i in c(5,7:9,14)){
  #all
  lac<- asa$lac_box
  var<- asa[,i]
  k<- colnames(asa)[i]
  all.corr<-cor.test(lac, var, method="pearson", alternative="two.sided")
  all.pcc<- round(all.corr$estimate,3)
  all.pval<- round(all.corr$p.value,3)
  #females
  lac.f<- fe$lac_box
  var.f<- fe[,i]
  fe.corr<-cor.test(lac.f, var.f, method="pearson", alternative="two.sided")
  fe.pcc<- round(fe.corr$estimate,3)
  fe.pval<- round(fe.corr$p.value,3)
  #males
  lac.m<- ma$lac_box
  var.m<- ma[,i]
  ma.corr<-cor.test(lac.m, var.m, method="pearson", alternative="two.sided")
  ma.pcc<- round(ma.corr$estimate,3)
  ma.pval<- round(ma.corr$p.value,3)
  #put
  put(k, all.pcc, fe.pcc, ma.pcc,
      all.pval, fe.pval, ma.pval)
}

cor_res<- magic_result_as_dataframe(F)
cor_res<- cor_res %>% tibble::column_to_rownames(var="k")

pcc<- cor_res %>% dplyr::select(all.pcc:ma.pcc)
pval<- cor_res %>% dplyr::select(all.pval:ma.pval)
colnames(pcc)<-c("All (n = 285)", "Women (n = 154)", "Men (n = 131)")
rownames(pcc)<- c("Age", "Height (cm)", "Weight (kg)", "BMI", "Kcal")

#png(filename="../plots/asa24/corrplot_lactose_metadata_all.png", units="in", height=4, width=5, res=300)
corrplot::corrplot(t(pcc), is.corr=F, tl.srt=65, p.mat=t(pval), sig.level = 0.05, insig="pch",
                   addCoef.col = "black",
                   title = "            Metadata and Lactose Consumption\nPearson's correlation",
                   mar=c(0,0,4,1))
#dev.off()
```

```{r scatterplot-kcalmeta}
for(i in c(5,7:9)){
  #filepath=paste("/Users/elizabeth.chin/Desktop/milk/DXA/plots_102419/genotype_sex/","genotypesex_vs_",colnames(cd)[i],".png",sep="")
  k<- colnames(asa[i])
  d<-data.frame(Subject=factor(asa$SubjectID),
                Genotype=asa$LCT,
                var=asa[,i],
                kcal=asa$kcal,
                Sex=factor(asa$sex))
  q<- ggplot(d, aes(x= var, y= kcal)) + 
    geom_point(aes(color=Genotype), size=3)+
    geom_sasath(method= "lm", se=F) +
    theme_bw() + xlab(print(k)) + 
    ylab("Kcal") +
    facet_wrap(~Sex, scales="free_x")+
    ggtitle(paste0(k," vs Total Intake (Kcal)")) +
    #stat_summary(fun.y=mean, geom="point", shape=17, size=4)+
    theme(plot.title = element_text(size=22),
          axis.title.x = element_text(size=22), 
          axis.title.y = element_text(size=22))
  print(q)
  #ggsave(file=filepath, plot=q, width=10, height=10)
}
```

Boxplot of height, age, and lactose intake by LCT genotype

```{r boxplot-genotype}
ggplot(data=asa, aes(x= age, y = ht_cm)) + geom_point(aes(color=LCT)) + theme_bw() + facet_wrap(~sex)
ggplot(data = asa, aes(x= LCT, y = ht_cm)) + geom_boxplot(aes(fill=LCT)) + facet_wrap(~sex)
ggplot(data=asa, aes(x = LCT, y = Lactose.consumed)) + geom_boxplot(aes(fill=LCT)) + facet_wrap(~sex)
```



```{r summaryse}
## Summarizes data.
## Gives count, mean, standard deviation, standard error of the mean, and confidence 
## interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE, conf.interval=.95) {
    library(doBy)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # Collapse the data
    formula <- as.formula(paste(measurevar, paste(groupvars, collapse=" + "), sep=" ~ "))
    datac <- summaryBy(formula, data=data, FUN=c(length2,mean,sd), na.rm=na.rm)

    # Rename columns
    names(datac)[ names(datac) == paste(measurevar, ".mean",    sep="") ] <- measurevar
    names(datac)[ names(datac) == paste(measurevar, ".sd",      sep="") ] <- "sd"
    names(datac)[ names(datac) == paste(measurevar, ".length2", sep="") ] <- "N"
    
    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
    
    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult
    
    return(datac)
}
```

Barplots of lactose and total dairy intake for ALL SUBJECTS

```{r barplots-genotype}
df<- summarySE(data=asa, measurevar="Lactose.consumed", groupvars="LCT")
ggplot(data=df, aes(x= LCT, y = Lactose.consumed)) + geom_bar(aes(fill = LCT), stat="identity") + 
  geom_errorbar(aes(ymax = Lactose.consumed + se, ymin = Lactose.consumed-se), width=0.2) + 
  theme_bw() + ggtitle("Average lactose consumed +/- SE")

df2<- summarySE(data=asa, measurevar="d_total", groupvars="LCT")
ggplot(data=df2, aes(x= LCT, y = d_total)) + geom_bar(aes(fill = LCT), stat="identity") + 
  geom_errorbar(aes(ymax = d_total + se, ymin = d_total-se), width=0.2) + 
  theme_bw() + ggtitle("Average Dairy Servings +/- SE")

df3<- summarySE(data=asa, measurevar="lac.kcal", groupvars="LCT")
ggplot(data=df3, aes(x= LCT, y = lac.kcal)) + geom_bar(aes(fill = LCT), stat="identity") + 
  geom_errorbar(aes(ymax = lac.kcal + se, ymin = lac.kcal-se), width=0.2) + 
  theme_bw() + ggtitle("Average lactose/kcal consumed +/- SE")

df4<- summarySE(data=asa, measurevar="dairy.kcal", groupvars="LCT")
ggplot(data=df4, aes(x= LCT, y = dairy.kcal)) + geom_bar(aes(fill = LCT), stat="identity") + 
  geom_errorbar(aes(ymax = dairy.kcal + se, ymin = dairy.kcal-se), width=0.2) + 
  theme_bw() + ggtitle("Average Dairy/KCal +/- SE")
```

check if lactose consumption varies with ethnicity (determines if we need to exclude some ethnicities)  

all ethnicities (including multi-racial subjects)
```{r boxplot-lactose-eth}
ggplot(data=asa, aes(x = Ethnicity, y = Lactose.consumed)) + geom_boxplot(aes(fill=Ethnicity),varwidth = T) + theme_bw()
ggplot(data=asa, aes(x = Ethnicity, y = d_total)) + geom_boxplot(aes(fill=Ethnicity),varwidth=T) + theme_bw()

Anova(fit<- lm(lac_box ~ kcal_box + Ethnicity, data=asa)) #0.01036
plot(fit)
Anova(fit2<- lm(total_tukey ~ kcal_box+ Ethnicity, data=asa)) #.03193
plot(fit2)

library(emmeans)
pairs(emm<- emmeans(fit2, "Ethnicity"))
#contrast                     estimate    SE  df t.ratio p.value
# Asian - Caucasian             -0.2575 0.0801 278 -3.215  0.0181

```

exclude multi-racial and & Other subjects  
```{r lm-eth}
dat<- asa %>% filter(Ethnicity != "Other" & Ethnicity != "Multi") #259 subjects
Anova(fit<- lm(lac_box ~ kcal_box + weightv2+ Ethnicity, data=dat)) #0.003846
plot(fit)
pairs(emmeans(fit, "Ethnicity"))
#Asian - Caucasian            -0.94939 0.318 253 -2.990  0.0161 

Anova(fit2<- lm(lac_box ~ Ethnicity, data=dat)) #0.001782
pairs(emmeans(fit2, "Ethnicity"))
# Asian - Caucasian              -1.217 0.324 255 -3.760  0.0012 
```

Does total dairy consumption differ between ethnicities?

```{r dtotal-eth}
Anova(fit<- lm(total_tukey ~ kcal_box + weightv2+ Ethnicity, data=dat)) #0.0249800 *
plot(fit)
pairs(emmeans(fit, "Ethnicity"))
#Asian - Caucasian             -0.1928 0.0790 253 -2.441  0.0721

Anova(fit2<- lm(total_tukey ~ Ethnicity, data=dat)) #0.00262 **
pairs(emmeans(fit2, "Ethnicity"))
#Asian - Caucasian             -0.3243 0.0874 255 -3.710  0.0014 
```


what about if we just use Caucasians and Hispanics?   
```{r make-ch}
ch<- asa %>% filter(Ethnicity == "Caucasian" | Ethnicity == "Hispanic") #n=215
summary(ch)
```

get the summary for each genotype

```{r summary-ch}
a<- aggregate(cbind(d_total, Lactose.consumed) ~ LCT, data=ch, FUN="mean")
b<- aggregate(cbind(d_total, Lactose.consumed) ~ LCT, data=ch, FUN="median")
c<- aggregate(cbind(d_total, Lactose.consumed) ~ LCT, data=ch, FUN="min")
d<- aggregate(cbind(d_total, Lactose.consumed) ~ LCT, data=ch, FUN="max")

colnames(a)<-c("LCT", "mean_dairy", "mean_lactose")
colnames(b)<-c("LCT", "median_dairy", "median_lactose")
colnames(c)<-c("LCT", "min_dairy", "min_lactose")
colnames(d)<-c("LCT", "max_dairy", "max_lactose")
df<- merge(merge(merge(a, b, by = "LCT"),c, by = "LCT"), d, by = "LCT")
htmlTable::htmlTable(df)

ch %>% group_by("LCT") %>% do(data.frame(summary(.)))


aa<- ch %>% filter(LCT == "AA") %>% dplyr::select(d_total, Lactose.consumed) %>% summary(.) 
ag<- ch %>% filter(LCT == "AG") %>% dplyr::select(d_total, Lactose.consumed) %>% summary(.)
gg<- ch %>% filter(LCT == "GG") %>% dplyr::select(d_total, Lactose.consumed) %>% summary(.)

cbind(aa, ag, gg) %>% htmlTable::htmlTable()

ggplot(data=ch, aes(x= LCT, y = Lactose.consumed)) + geom_boxplot(aes(fill=LCT)) + geom_jitter(width=.25, col= "blue")

#how many LNP had lactose consumption > 20? 
ch %>% filter(Lactose.consumed >= 12) %>% plyr::count(., "LCT") #16 GG
ch %>% filter(Lactose.consumed >= 15) %>% plyr::count(., "LCT") #7 GG
ch %>% filter(Lactose.consumed >= 20) %>% plyr::count(., "LCT") #2 GG

#total n caucasians and hispanics per genotype
plyr::count(ch, "LCT")
#AA: 62
#AG: 92
#GG: 61 
```

correlation of d_total and lactose among caucasians and hispanics
```{r cor-ch}
cor.test(ch$total_tukey, ch$lac_box, method="pearson", alternative="two.sided")
```

```{r ch-lactosevdtotal}
p<-ggplot(data=ch, aes(x= d_total, y = Lactose.consumed)) + geom_point(aes(color=LCT), size = 2) + theme_bw() + 
  ggtitle("Average Lactose Consumption vs Dairy Intake ") + ylab("Estimated Lactose Consumption (g)") +
  xlab("Total Dairy Consumption (servings)") + geom_sasath(method="lm", se=F, col= "darkgray", linetype="dashed") + 
  theme(legend.position = "bottom", 
        plot.title = element_text(hjust=0.5))
ggsave(p,file = "../plots/asa24/lactose_vs_dtotal_ch.png", height=4, width=5, units="in", dpi=300)
```

```{r ch-jitter}
for.plot<-ch
library(plotly)
shapes<- c("circle", "cross")
for.plot$shapes<- shapes[for.plot$sex]
for.plot$col<- rainbow(3)[for.plot$LCT]

plot_ly(data=ch, x = ~Lactose.consumed, type="scatter",
        text=~paste("Subject:",SubjectID,
                    "<br>Ethnicity:", Ethnicity,
                    "<br>Age:", age,
                    "<br>BMI:", bmi_final),
        color=~LCT, 
        symbol= ~sex,
        size=3) %>% 
  layout(xaxis=list(
    title="Average Lactose Consumed (g)"), 
    yaxis= list(
      title="Random Jitter"), 
    title="Average Lactose Consumed for \nCaucasian and Hispanic Subjects (n = 215)"
  )

```

```{r scatterplot-kcal-metadata-ch}
for(i in c(5,7:9)){
  #filepath=paste("/Users/elizabeth.chin/Desktop/milk/DXA/plots_102419/genotype_sex/","genotypesex_vs_",colnames(cd)[i],".png",sep="")
  k<- colnames(ch[i])
  d<-data.frame(Subject=factor(ch$SubjectID),
                Genotype=ch$LCT,
                var=ch[,i],
                kcal=ch$kcal,
                Sex=factor(ch$sex))
  q<- ggplot(d, aes(x= var, y= kcal)) + 
    geom_point(aes(color=Genotype), size=3)+
    geom_sasath(method= "lm", se=F) +
    theme_bw() + xlab(print(k)) + 
    ylab("Kcal") +
    facet_wrap(~Sex, scales="free_x")+
    ggtitle(paste0(k," vs Total Intake (Kcal) in Caucasians and Hispanics")) +
    #stat_summary(fun.y=mean, geom="point", shape=17, size=4)+
    theme(plot.title = element_text(size=22),
          axis.title.x = element_text(size=22), 
          axis.title.y = element_text(size=22))
  print(q)
  #ggsave(file=filepath, plot=q, width=10, height=10)
}
```


```{r lm-eth-ch}

Anova(fit<- lm(lac_box ~ kcal_box + Ethnicity, data=ch)) #0.2617
plot(fit)
Anova(fit2<- lm(kcal_box ~ Ethnicity, data=ch)) #0.8963
plot(fit2)
ggplot(data=ch, aes(x = Ethnicity, y = Lactose.consumed)) + geom_boxplot(aes(fill=Ethnicity),varwidth = T) + theme_bw()
ggplot(data=ch, aes(x = kcal, y = Lactose.consumed)) + geom_point(aes(color=Ethnicity)) + theme_bw()

```

```{r cortest-rawvalues-ch}
for(i in c(5,7:9,14)){
  k<- colnames(ch)[i]
  var<- ch[,i]
  lac<- ch$lac_box
  c<-cor.test(lac, var, alternative="two.sided", method="pearson")
  print(k)
  print(c)
}

#women only
fe<- ch %>% filter(sex=="female")
for(i in c(5,7:9,14)){
  k<- colnames(fe)[i]
  var<- fe[,i]
  lac<- fe$lac_box
  c<-cor.test(lac, var, alternative="two.sided", method="pearson")
  print(k)
  print(c)
}

ma<- ch %>% filter(sex=="male")
for(i in c(5,7:9,14)){
  k<- colnames(ma)[i]
  var<- ma[,i]
  lac<- ma$lac_box
  c<-cor.test(lac, var, alternative="two.sided", method="pearson")
  print(k)
  print(c)
}

#just to be sure
Anova(fit<-lm(kcal_box ~ age, data=ch)) #0.5881
plot(fit)
```




```{r corrplot-lactose-meta-ch}
library(magicfor)
magic_for()
fe<- ch %>% filter(sex == "female")
ma<- ch %>% filter(sex=="male")
for(i in c(5,7:9,14)){
  #all
  lac<- ch$lac_box
  var<- ch[,i]
  k<- colnames(ch)[i]
  all.corr<-cor.test(lac, var, method="pearson", alternative="two.sided")
  all.pcc<- round(all.corr$estimate,3)
  all.pval<- round(all.corr$p.value,3)
  #females
  lac.f<- fe$lac_box
  var.f<- fe[,i]
  fe.corr<-cor.test(lac.f, var.f, method="pearson", alternative="two.sided")
  fe.pcc<- round(fe.corr$estimate,3)
  fe.pval<- round(fe.corr$p.value,3)
  #males
  lac.m<- ma$lac_box
  var.m<- ma[,i]
  ma.corr<-cor.test(lac.m, var.m, method="pearson", alternative="two.sided")
  ma.pcc<- round(ma.corr$estimate,3)
  ma.pval<- round(ma.corr$p.value,3)
  #put
  put(k, all.pcc, fe.pcc, ma.pcc,
      all.pval, fe.pval, ma.pval)
}

cor_res<- magic_result_as_dataframe(F)
cor_res<- cor_res %>% tibble::column_to_rownames(var="k")

pcc<- cor_res %>% dplyr::select(all.pcc:ma.pcc)
pval<- cor_res %>% dplyr::select(all.pval:ma.pval)
colnames(pcc)<-c("All (n = 215)", "Women (n = 117)", "Men (n = 98)")
rownames(pcc)<- c("Age", "Height (cm)", "Weight (kg)", "BMI", "Kcal")

png(filename="../plots/asa24/corrplot_lactose_metadata.png", units="in", height=4, width=5, res=300)
corrplot::corrplot(t(pcc), is.corr=F, tl.srt=65, p.mat=t(pval), sig.level = 0.05, insig="pch",
                   addCoef.col = "black",
                   title = "            Caucasian and Hispanic\nMetadata and Lactose Consumption\nPearson's correlation",
                   mar=c(0,0,4,1))
dev.off()
```



```{r corrplot-kcal-meta-ch}
magic_for()
for(i in c(5,7:9)){
  #all
  kcal<- ch$kcal_box
  var<- ch[,i]
  k<- colnames(ch)[i]
  all.corr<-cor.test(kcal, var, method="pearson", alternative="two.sided")
  all.pcc<- round(all.corr$estimate,3)
  all.pval<- round(all.corr$p.value,3)
  #females
  kcal.f<- fe$kcal_box
  var.f<- fe[,i]
  fe.corr<-cor.test(kcal.f, var.f, method="pearson", alternative="two.sided")
  fe.pcc<- round(fe.corr$estimate,3)
  fe.pval<- round(fe.corr$p.value,3)
  #males
  kcal.m<- ma$kcal_box
  var.m<- ma[,i]
  ma.corr<-cor.test(kcal.m, var.m, method="pearson", alternative="two.sided")
  ma.pcc<- round(ma.corr$estimate,3)
  ma.pval<- round(ma.corr$p.value,3)
  #put
  put(k, all.pcc, fe.pcc, ma.pcc,
      all.pval, fe.pval, ma.pval)
}

cor_res<- magic_result_as_dataframe(F)
cor_res<- cor_res %>% tibble::column_to_rownames(var="k")

pcc<- cor_res %>% dplyr::select(all.pcc:ma.pcc)
pval<- cor_res %>% dplyr::select(all.pval:ma.pval)
colnames(pcc)<-c("All (n = 215)", "Women (n = 117)", "Men (n = 98)")
rownames(pcc)<- c("Age", "Height (cm)", "Weight (kg)", "BMI")

png(filename="../plots/asa24/corrplot_kcal_metadata.png", units="in", height=4, width=5, res=300)
corrplot::corrplot(t(pcc), is.corr=F, tl.srt=65, p.mat=t(pval), sig.level = 0.05, insig="pch",
                   addCoef.col = "black",
                   title = "            Caucasian and Hispanic\nMetadata and Total KCal\nPearson's correlation",
                   mar=c(0,0,4,1))
dev.off()
```
```{r corrplot-dtotal}
magic_for()
for(i in c(5,7:9,14)){
  #all
  dt<- ch$total_tukey
  var<- ch[,i]
  k<- colnames(ch)[i]
  all.corr<-cor.test(dt, var, method="pearson", alternative="two.sided")
  all.pcc<- round(all.corr$estimate,3)
  all.pval<- round(all.corr$p.value,3)
  #females
  dt.f<- fe$total_tukey
  var.f<- fe[,i]
  fe.corr<-cor.test(dt.f, var.f, method="pearson", alternative="two.sided")
  fe.pcc<- round(fe.corr$estimate,3)
  fe.pval<- round(fe.corr$p.value,3)
  #males
  dt.m<- ma$total_tukey
  var.m<- ma[,i]
  ma.corr<-cor.test(dt.m, var.m, method="pearson", alternative="two.sided")
  ma.pcc<- round(ma.corr$estimate,3)
  ma.pval<- round(ma.corr$p.value,3)
  #put
  put(k, all.pcc, fe.pcc, ma.pcc,
      all.pval, fe.pval, ma.pval)
}

cor_res<- magic_result_as_dataframe(F)
cor_res<- cor_res %>% tibble::column_to_rownames(var="k")

pcc<- cor_res %>% dplyr::select(all.pcc:ma.pcc)
pval<- cor_res %>% dplyr::select(all.pval:ma.pval)
colnames(pcc)<-c("All (n = 215)", "Women (n = 117)", "Men (n = 98)")
rownames(pcc)<- c("Age", "Height (cm)", "Weight (kg)", "BMI", "Kcal")

png(filename="../plots/asa24/corrplot_dtotal_metadata.png", units="in", height=4, width=5, res=300)
corrplot::corrplot(t(pcc), is.corr=F, tl.srt=65, p.mat=t(pval), sig.level = 0.05, insig="pch",
                   addCoef.col = "black",
                   title = "            Caucasian and Hispanic\nMetadata and Total Dairy\nPearson's correlation",
                   mar=c(0,0,4,1))
dev.off()
```


Do any of the metadata significantly differ among Genotypes?   
Height significantly differs among genotypes  
```{r}
for(i in c(5,7:9)){
  k<- colnames(ch)[i]
  var<- ch[,i]
  fit<- kruskal.test(var ~ LCT, data=ch)
  print(k)
  print(fit)
  #plot(fit, main=paste(k))
  #plot(fit$residuals, main=paste(k))
}

pairwise.wilcox.test(ch$ht_cm, ch$LCT, p.adjust.method="BH")
ggplot(data=ch, aes(x=LCT, y = ht_cm)) + geom_boxplot() #AA and AG subjects are significantly taller than GG subjects

#linear
Anova(fit<- lm(log(ht_cm) ~ LCT, data=ch))
plot(fit)
hist(fit$residuals)
Anova(fit2<- lm(log(ht_cm) ~ kcal_box + lac_box, data=ch))
#mediation
Anova(fit3<- lm(log(ht_cm) ~ lac_box + LCT, data=ch ))
```


does lactose consumption differ among genotypes   

all subjects
```{r lm-lactose-lct}

#all subjects, base model
Anova(fit<-lm(lac_box ~ LCT, data=asa))
#LCT        49.19   2  8.4103 0.0002833 ***
plot(fit)
Anova(fit1<-lm(lac_box ~ LCT*sex, data=asa)) #no significant interaction
#LCT        45.97   2  7.9695 0.0004307 ***
#sex         7.62   1  2.6440 0.1050696    
#LCT:sex    12.53   2  2.1717 0.1159046    
Anova(fit2<-lm(lac_box ~ sex + LCT, data=asa))
#sex         7.62   1  2.6221 0.1065036    
#LCT        45.97   2  7.9036 0.0004577 ***

#all  single-ethnicity subjects, base with ethnicity and genotype
Anova(fit.base<-lm(lac_box ~ Ethnicity + LCT, data=dat))
#        Sum Sq  Df F value  Pr(>F)  
#Ethnicity  16.08   3  1.8441 0.13963  
#LCT        24.17   2  4.1576 0.01673 *

#full model for single-ethnicity subjects: 
Anova(fit5<-lm(lac_box ~ kcal_box + log(weightv2)+Ethnicity + LCT, data=dat))
#kcal_box       51.71   1 19.3854 1.582e-05 ***
#log(weightv2)   3.66   1  1.3724    0.2425    
#Ethnicity      12.67   3  1.5833    0.1939    
#LCT            15.36   2  2.8798    0.0580 .  
plot(fit5)
hist(fit5$residuals)


#all subjects, full model:
Anova(fit8<-lm(lac_box ~ kcal_box + log(weightv2) + LCT, data=asa)) #0.003475
plot(fit8)
hist(fit8$residuals)
pairs(emmeans(fit8, "LCT"))
# contrast estimate    SE  df t.ratio p.value
# AA - AG     0.558 0.256 280 2.182   0.0760 
# AA - GG     0.864 0.254 280 3.399   0.0022 
# AG - GG     0.306 0.221 280 1.388   0.3488  
Anova(fita<-lm(lac_box ~ LCT, data=asa))
pairs(emmeans(fita, "LCT"))
```

all subjects- subset by sex, lactose

```{r}
fe.all<- asa %>% filter(sex == "female")
ma.all<- asa %>% filter(sex == "male")
#all females, full model
Anova(fit.f<- lm(lac_box ~ kcal_box + log(weightv2) + LCT, data=fe.all)) #0.362850  
plot(fit.f)
hist(fit.f$residuals)

#all males, full model
Anova(fit.m<- lm(lac_box ~ kcal_box + log(weightv2) + LCT, data=ma.all)) # 0.003131
plot(fit.m)
pairs(emmeans(fit.m, "LCT"))
# AA - AG     0.992 0.387 126 2.561   0.0310 
# AA - GG     1.330 0.392 126 3.394   0.0026 
```


caucasian only- lactose
```{r}
#caucasian and hispanics, basemodel 
Anova(fit.basech<-lm(lac_box ~ LCT, data=ch)) #0.006989 **
pairs(emmeans(fit.basech, "LCT"))
# AA - GG     0.926 0.294 212 3.154   0.0052 

#there is no interaction with sex
Anova(mod<- lm(lac_box ~ kcal_box + sex*LCT, data=ch))

#caucasian and hispanics, full model: 
Anova(fit2<- lm(lac_box ~ log(weightv2) + kcal_box + LCT, data=ch))
#log(weightv2)   2.26   1  0.9407   0.33322    
#kcal_box       47.08   1 19.5998 1.535e-05 ***
#LCT            18.01   2  3.7491   0.02513 *  
plot(fit2)
hist(fit2$residuals)
emmeans(fit2, "LCT")
#LCT emmean    SE  df lower.CL upper.CL
# AA    3.59 0.197 210     3.20     3.98
# AG    3.07 0.162 210     2.75     3.39
# GG    2.84 0.202 210     2.45     3.24
pairs(emmeans(fit2, "LCT"))
# contrast estimate    SE  df t.ratio p.value
# AA - AG     0.520 0.255 210 2.039   0.1055 
# AA - GG     0.749 0.283 210 2.644   0.0238 
# AG - GG     0.228 0.258 210 0.886   0.6497 
```

Caucasian men and women subsets- lactose
```{r}
fe<- ch %>% filter(sex == "female")
ma<- ch %>% filter(sex == "male")
#women
Anova(fit3<- lm(lac_box ~ log(weightv2) + kcal_box + LCT, data=fe)) #no significant differences
#log(weightv2)   7.562   1  3.7276 0.05605 .
#kcal_box        4.714   1  2.3236 0.13024  
#LCT             3.323   2  0.8190 0.44351  
#Residuals     227.211 112   
plot(fit3)
hist(fit3$residuals)

#men
Anova(fit4<- lm(lac_box ~ log(weightv2) + kcal_box + LCT, data=ma))
#log(weightv2)   0.149  1  0.0554   0.8144    
#kcal_box       53.246  1 19.8423 2.34e-05 ***
#LCT            24.254  2  4.5191   0.0134 * 
plot(fit4)
hist(fit4$residuals)
pairs(emmeans(fit4, "LCT"))
# AA - AG     0.995 0.381 93 2.609   0.0283 
# AA - GG     1.151 0.450 93 2.558   0.0323 
# AG - GG     0.156 0.429 93 0.363   0.9301 

```


Does genotype influence total dairy intake (servings)?   
```{r}
#all subjects, base model
Anova(fit2<- lm(total_tukey ~ LCT, data=asa)) #0.007869
plot(fit2)
pairs(emmeans(fit2, "LCT")) #AA - GG     0.229 0.0738 282 3.098   0.0061
Anova(fit2<- lm(total_tukey ~ LCT*sex, data=asa)) #no significant interaction
Anova(fit2<- lm(total_tukey ~ sex + LCT, data=asa)) 

#all subjects, full model
Anova(fit3<- lm(total_tukey ~ log(weightv2) + kcal_box + LCT, data=asa)) #0.1393189
plot(fit3)
hist(fit3$residuals)

#all single-eth subjects, full model
Anova(fit.all<- lm(total_tukey ~ log(weightv2) + kcal_box + Ethnicity + LCT, data=dat)) #0.6399354
plot(fit.all)
hist(fit.all$residuals)

#caucasian and hispanics only, full model
Anova(fit.ch<- lm(total_tukey ~ log(weightv2) + kcal_box + LCT, data=ch)) #0.5867361 

#Caucasian and hispanics only, base model
Anova(fit.chb<- lm(total_tukey ~ LCT, data=ch)) #0.155 

```

total dairy- subset by sex: 
```{r}
#females, all
Anova(fit.fall<- lm(total_tukey ~ kcal_box + log(weightv2) + LCT, data=fe.all)) #0.6831004

#females, ch
Anova(fit.fch<- lm(total_tukey ~ kcal_box + log(weightv2) + LCT, data=fe)) #0.7067424

#males, all
Anova(fit.mall<- lm(total_tukey ~ kcal_box + log(weightv2) + LCT, data=ma.all)) #0.15783

#males, ch
Anova(fit.mch<- lm(total_tukey ~ kcal_box + log(weightv2) + LCT, data=ma)) #0.44590

plot(fit.fall)
hist(fit.fall$residuals, breaks=10)

plot(fit.fch)
hist(fit.fch$residuals, breaks=10)

plot(fit.mall)
hist(fit.mall$residuals, breaks=10)

plot(fit.mch)
hist(fit.mch$residuals, breaks=10)
```



there are similar results using transformed variables corresponding  to lactose/kcal  and totaldairy/kcal, but I think it's a little easier to interpret/explain an adjusted model instead of a normalized variable.   
```{r echo=F, eval=F}
Anova(mod<- lm(lackcal_box ~ LCT, data=ch))
plot(mod)
hist(mod$residuals)
Anova(mod2<- lm(dairykcal_tukey ~ LCT, data=ch))
```

```{r}
ggplot(data=asa, aes(x = LCT, y = kcal)) + geom_boxplot()
Anova(lm(kcal_box ~ LCT, data=asa))
ggplot(data=asa, aes(x = LNP, y = kcal)) + geom_boxplot()
Anova(lm(kcal_box ~ LNP, data=asa))
```

more plots   
```{r barplot-ch-genotype}
df<- summarySE(data=ch, measurevar="Lactose.consumed", groupvars="LCT")
ggplot(data=df, aes(x= LCT, y = Lactose.consumed)) + geom_bar(aes(fill = LCT), stat="identity") + 
  geom_errorbar(aes(ymax = Lactose.consumed + se, ymin = Lactose.consumed-se), width=0.2) + 
  theme_bw() + ggtitle("Average lactose consumed by LCT Genotype") +
  xlab("LCT Genotype") + ylab("Estimated Lactose Consumed +/- SE (g)") +
  theme(legend.position="none", 
        plot.title = element_text(hjust=0.5))

ggsave("../plots/asa24/lactose_lct_barplot.png", width=5, height=4, units="in", dpi=300)

df2<- summarySE(data=ch, measurevar="Lactose.consumed", groupvars=c("sex", "LCT"))
ggplot(data=df2, aes(x= LCT, y = Lactose.consumed)) + geom_bar(aes(fill = LCT), stat="identity") + 
  geom_errorbar(aes(ymax = Lactose.consumed + se, ymin = Lactose.consumed-se), width=0.2) + 
  facet_wrap(~sex)+
  theme_bw() + ggtitle("Average lactose consumed by LCT Genotype") +
  xlab("LCT Genotype") + ylab("Estimated Lactose Consumed +/- SE (g)") +
  theme(legend.position="none", 
        plot.title = element_text(hjust=0.5))

ggsave("../plots/asa24/lactose_lct_barplot_faceted.png", width=5, height=4, units="in", dpi=300)

df3<- summarySE(data=ch, measurevar="d_total", groupvars="LCT")
ggplot(data=df3, aes(x= LCT, y = d_total)) + geom_bar(aes(fill = LCT), stat="identity") + 
  geom_errorbar(aes(ymax = d_total + se, ymin = d_total-se), width=0.2) + 
  theme_bw() + ggtitle("Average Dairy Servings by LCT Genotype") +
  xlab("LCT Genotype") + ylab("Dairy Consumption +/- SE (servings)") +
  theme(legend.position="none", 
        plot.title = element_text(hjust=0.5))
ggsave("../plots/asa24/dtotal_lct_barplot.png", width=5, height=4, units="in", dpi=300)


```

```{r barplot-ethnicity}
df4<- summarySE(data=dat, measurevar="Lactose.consumed", groupvars="Ethnicity")
ggplot(data=df4, aes(x= Ethnicity, y = Lactose.consumed)) + geom_bar(aes(fill = Ethnicity), stat="identity") + 
  geom_errorbar(aes(ymax = Lactose.consumed + se, ymin = Lactose.consumed-se), width=0.2) + 
  theme_bw() + ggtitle("Lactose Consumed by Ethnicity") +
  xlab("Ethnicity") + ylab("Estimated Lactose +/- SE (g)") +
  theme(legend.position="none", 
        plot.title = element_text(hjust=0.5))
ggsave("../plots/asa24/lactose_ethnicity_barplot.png", width=5, height=4, units="in", dpi=300)

df5<- summarySE(data=dat, measurevar="d_total", groupvars="Ethnicity")
ggplot(data=df5, aes(x= Ethnicity, y = d_total)) + geom_bar(aes(fill = Ethnicity), stat="identity") + 
  geom_errorbar(aes(ymax = d_total + se, ymin = d_total-se), width=0.2) + 
  theme_bw() + ggtitle("Average Dairy Servings by Ethnicity") +
  xlab("Ethnicity") + ylab("Dairy Consumption +/- SE (servings)") +
  theme(legend.position="none", 
        plot.title = element_text(hjust=0.5))
```

