---
title: "02_ASA24_LNP_analysis_caucasians_hispanics"
---

Author: E.Chin
Purpose: ASA24 Lactose and D_TOTAL analysis w.r.t. LNP genotypes; for use with 16s analysis paper. This is for only the Caucasian and Hispanic subjects. 


```{r setup}
library(dplyr)
library(ggplot2)
library(car)
library(doBy)
library(GGally)
library(corrplot)
library(cowplot)
library(emmeans)
library(MASS)
library(rcompanion)

asa<- read.csv("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/out_asa24/03_asa24_lactose_dtotal_no_powders_lct.csv", header=T) #these are the subjects who have lactose (average over 2 or 3 recalls) estimates available from usable ASA24 recalls  

#Only keep the Caucasian and Hispanic subjects
asa<-asa %>% filter(Ethnicity == "Caucasian" | Ethnicity == "Hispanic")
asa$Ethnicity<-as.factor(asa$Ethnicity)
asa$LCT<-as.factor(asa$LCT)
asa$LNP<-as.factor(asa$LNP)


male<- asa %>% filter(sex == 1)
female<- asa %>% filter(sex == 2)
```

```{r}
summary(male)
summary(female)
```

```{r}
plyr::count(asa, c("sex", "LNP")) %>% htmlTable::htmlTable(rnames = F)
plyr::count(asa, c("sex", "LCT")) %>% htmlTable::htmlTable(rnames=F)
```

```{r}
makecorrs<- function(data, titlename){
  asatmp<- data %>% dplyr::select(SubjectID, age, ht_cm:bmi_final, Lactose.consumed:KCAL)
  asacorr<- cor(asatmp[,-1])
  title = paste0("corrplot for ", titlename)
  asap<- ggcorrplot::cor_pmat(asatmp[,-1])
  cplot<- corrplot::corrplot(asacorr, is.corr=F, tl.srt=65, p.mat=asap, sig.level = 0.05, insig="pch",
                   mar=c(0,0,3,0), title = title)
  return(cplot)
}
```

```{r}
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/corrplot_asa_men.pdf", width = 4, height =4)
corr_men<- makecorrs(male, "men")
dev.off()

pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/corrplot_asa_women.pdf", width = 4, height =4)
corr_women<- makecorrs(female, "women")
dev.off()
```
```{r}
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/scatter_asa_lactose_dtotal_sex.pdf", width = 6, height =4)
ggplot(data = asa, aes(x = D_TOTAL, y = Lactose.consumed, color = LNP)) + geom_point(aes(fill =LNP), size = 2) + 
  geom_rug()+
  facet_wrap(~factor(sex), scales = "free") + theme_bw()
dev.off()
```
# Check out distributions

```{r}
hist(male$Lactose.consumed)
hist(male$D_TOTAL)
hist(female$Lactose.consumed)
hist(female$D_TOTAL)
```

```{r}
bc_transform<- function(data, var){
  for_trans<- data[,var]
  Box = boxcox(for_trans ~ 1,              # Transform Turbidity as a single vector
             lambda = seq(-6,6,0.1))      # Try values -6 to 6 by 0.1
  Cox = data.frame(Box$x, Box$y)            # Create a data frame with the results
  Cox2 = Cox[with(Cox, order(-Cox$Box.y)),] # Order the new data frame by decreasing y
  Cox2[1,]                                  # Display the lambda with the greatest
                                          #    log likelihood
  lambda = Cox2[1, "Box.x"]                 # Extract that lambda 0.3
  print(lambda)
  dat_box = (for_trans ^ lambda - 1)/lambda   # Transform the original data
  hist(dat_box)
  shapiro.test(dat_box) %>% print()
  newname<- paste0(var,"_box")
  data[,newname]<- dat_box
  return(data)
}
```


transform variables
```{r transform-lactose}
male<- bc_transform(data = male, var = "Lactose.consumed") #lambda = 0.3
female<- bc_transform(data = female, var = "Lactose.consumed") #lambda = 0.4

#in case we need to backtransform, let's use the same labmdas for kcal
male$kcal_box<- (male$KCAL ^ 0.3 - 1) / 0.3
female$kcal_box<- (female$KCAL ^ 0.4 - 1) / 0.3

hist(male$kcal_box)
hist(female$kcal_box)

#there are 0s in female$D_TOTAL, so we'll transform tukey
female$total_tukey<-transformTukey(female$D_TOTAL)
male$total_tukey<-transformTukey(male$D_TOTAL)


```

```{r}
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/boxplot_asa_lactose_lct_sex.pdf", width = 6, height =4)
ggplot(data = asa, aes(x= LCT, y = Lactose.consumed)) + geom_boxplot(aes(fill = LNP), varwidth = T, notch = T) + 
  theme_bw() + facet_wrap(~sex)
dev.off()
```


# Does Lactose consumption significantly differ between LP and LNP? 
```{r}
ix_plots<- function(data, varx, vary){
  plot_x<- data[,varx]
  plot_y <- data[,vary]
  q<- ggplot(data = data, aes(x = plot_x, y = plot_y, color = LNP)) + geom_point(aes(fill = LNP)) + 
    geom_smooth(method = "lm", se = F, aes(color = LNP)) + theme_bw() + xlab(varx) + ylab(vary)
  print(q)
}

ix_plots(data = male, varx = "KCAL", vary="Lactose.consumed")
ix_plots(data = male, varx = "KCAL", vary="D_TOTAL")
ix_plots(data = female, varx = "KCAL", vary="Lactose.consumed")
ix_plots(data = female, varx = "KCAL", vary="D_TOTAL")
```

Does D_TOTAL or Lactose consumption differ with LNP status in MEN?
```{r}
t.test(data = male, total_tukey ~ LNP, alternative = "two.sided") #0.4902
t.test(data = male, Lactose.consumed_box ~ LNP, alternative = "two.sided") #0.09469
```
```{r}
do_lm<- function(df, yvar, x_main, x_cov){
  y = df[,yvar]
  xvar = df[,x_main]
  xcov= df[,x_cov]
  fit_ix<- lm(y ~ xcov*xvar, data = df)
  s<- (summary(fit_ix))
  d<- s$coefficients %>% as.data.frame()
  if(d$`Pr(>|t|)`[4] < 0.05){
    plot(fit_ix)
    hist(fit_ix$residuals)
    print(s)
  }else if(d$`Pr(>|t|)`[4] >= 0.05){
    print("no significant interction")
    fit<- lm(y ~ xcov + xvar, data = df)
    plot(fit)
    hist(fit$residuals)
    car::Anova(fit) %>% print()
  }

}
```

```{r}
do_lm(df = male, yvar = "Lactose.consumed_box", x_main = "LNP", x_cov = "kcal_box")
```

```{r}
do_lm(df = male, yvar = "total_tukey", x_main = "LNP", x_cov = "kcal_box")
```

```{r}
t.test(data = female, total_tukey ~ LNP, alternative = "two.sided") #0.1608
t.test(data = female, Lactose.consumed_box ~ LNP, alternative = "two.sided") #0.1117
```


```{r}
do_lm(df = female, yvar = "Lactose.consumed_box", x_main = "LNP", x_cov = "kcal_box") #residuals are a little weird 

do_lm(df = female, yvar = "total_tukey", x_main = "LNP", x_cov = "kcal_box")
```

Does lactose or total dairy significantly differ among LCT genotypes in men and women?

```{r}
car::Anova(lm(total_tukey ~ LCT, data = male))
car::Anova(lm(Lactose.consumed_box ~ LCT, data = male))

do_lm(df = male, yvar = "total_tukey", x_main = "LCT", x_cov = "kcal_box")
do_lm(df = male, yvar = "Lactose.consumed_box", x_main = "LCT", x_cov = "kcal_box")

#residuals of lactose lm looks a little weird, let's do ordinal logistic rgression as a 'non parametric' way to check 
MASS::polr(factor(rank(Lactose.consumed))~ LCT, data = male) %>% car::Anova()
```
get marginal means for LCT/lactose in men 
```{r}
bctran <- make.tran("boxcox", 0.3)
lct_fit <- with(bctran, 
                lm(linkfun(Lactose.consumed) ~ linkfun(KCAL)+LCT, data = male))
car::Anova(lct_fit)
emmeans(lct_fit, pairwise~LCT, type = "response")
```


```{r}
car::Anova(lm(total_tukey ~ LCT, data = female))
car::Anova(lm(Lactose.consumed_box ~ LCT, data = female))

do_lm(df = female, yvar = "total_tukey", x_main = "LCT", x_cov = "kcal_box")
do_lm(df = female, yvar = "Lactose.consumed_box", x_main = "LCT", x_cov = "kcal_box")

```

