---
title: "04_compare_ffq_asa.Rmd"
output: 
  html_document:
    toc: true
    toc_float: true
---

Author: E.Chin
Purpose: Compare ASA and FFQ lactose. Will help guide how much FFQ analysis to do. 

```{r message=F, echo = F}
library(dplyr)
library(ggplot2)
library(plotly)
```

```{r, echo = F}
setwd("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/")
ffq<- read.csv("out_ffq/02_ffq_lactose_dtotal_lct.csv", header=T)
asa<-read.csv("out_asa24/03_asa24_lactose_dtotal_no_powders_lct.csv", header=T)
```

```{r, echo = F}
dim(ffq) #346 x 15
dim(asa) #286 x 15
colnames(ffq)
colnames(asa)
```
# Comparing ASA24 and FFQ data

283 subjects have both FFQ and ASA24 data 
```{r}
#who overlaps
gplots::venn(list("n FFQ subjects" = ffq$SubjectID, "n ASA subjects" = asa$SubjectID))
```


```{r echo = F}
#merge
df<- merge(ffq, asa, by = c("SubjectID", "LCT", "LNP", "Ethnicity", "age", "sex", "ht_cm", "weightv2", "bmi_final", "bin_number", "age_cat", "bmi_cat"), all =T)
colnames(df)
```
# Comparing Lactose

Plot of FFQ lactose (lacs) and ASA24 lactose (Lactose.consumed).     
Color corresponds to LP status, size with bmi_final, and shape corresponds to sex (1 = Male, 2 = Female). Hover over the dots for more metadata info.  
```{r warning=FALSE, message=FALSE}
plot_ly(df, x = ~lacs, y = ~Lactose.consumed, 
        color = ~LNP, symbol = ~factor(sex), size = ~bmi_final,
        text = ~paste("SubjectID: ", SubjectID,
                      "<br>Ethnicity: ", Ethnicity,
                      "<br>LCT: ", LCT))
```

```{r}
cor.test(df$lacs, df$Lactose.consumed, method = "pearson", alternative = "two.sided")
```

# Comparing total dairy

Plot of FFQ total dairy (d_total, x axis) and ASA24 total dairy (D_TOTAL, y axis).    
Color corresponds to LP status, size with bmi_final, and shape corresponds to sex (1 = Male, 2 = Female). Hover over the dots for more metadata info.  
```{r warning=FALSE, message=FALSE}
plot_ly(df, x = ~d_total, y = ~D_TOTAL, 
        color = ~LNP, symbol = ~factor(sex), size = ~bmi_final,
        text = ~paste("SubjectID: ", SubjectID,
                      "<br>Ethnicity: ", Ethnicity,
                      "<br>LCT: ", LCT))
```

```{r}
cor.test(df$d_total, df$D_TOTAL, method = "pearson", alternative = "two.sided")
```


# Compare total caloric intake

Plot of FFQ total intake (dt_kcal, x axis) and ASA24 total intake (KCAL, y axis).    
Color corresponds to LP status, size with bmi_final, and shape corresponds to sex (1 = Male, 2 = Female). Hover over the dots for more metadata info.  
```{r warning=FALSE, message=FALSE}
plot_ly(df, x = ~dt_kcal, y = ~KCAL, 
        color = ~LNP, symbol = ~factor(sex), size = ~bmi_final,
        text = ~paste("SubjectID: ", SubjectID,
                      "<br>Ethnicity: ", Ethnicity,
                      "<br>LCT: ", LCT))
```

```{r}
cor.test(df$dt_kcal, df$KCAL, method = "pearson", alternative = "two.sided")
```

# Compare FFQ total dairy to FFQ lactose

Plot of FFQ total dairy (d_total, x axis) and FFQ lactose (lacs, y axis).    
Color corresponds to LP status, size with bmi_final, and shape corresponds to sex (1 = Male, 2 = Female). Hover over the dots for more metadata info.  

```{r warning=FALSE, message=FALSE}
plot_ly(df, x = ~d_total, y = ~lacs, 
        color = ~LNP, symbol = ~factor(sex), size = ~bmi_final,
        text = ~paste("SubjectID: ", SubjectID,
                      "<br>Ethnicity: ", Ethnicity,
                      "<br>LCT: ", LCT))

cor.test(df$d_total, df$lacs, method = "pearson", alternative = "two.sided")
```

