---
title: "05_ffq_lct_analysis"
---

Author: E.Chin
Purpose: FFQ Lactose and d_total analysis w.r.t. LNP genotypes; for use with 16s analysis paper. 

Note 02/11/2022- we will exclude LNP african americans from all-ethnicity analysis since we cannot be sure of their LP status. Multi-ethnic African American/Black participants are 6083 and 8015 

```{r setup}
library(dplyr)
library(ggplot2)
library(car)
library(doBy)
library(GGally)
library(corrplot)
library(cowplot)
library(emmeans)

ffq<- read.csv("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/out_ffq/02_ffq_lactose_dtotal_lct.csv", header=T) #these are the subjects who have lactose (average over 2 or 3 recalls) estimates available from usable ffq24 recalls  

#remove African American subjects who are GG
af_gg<- ffq %>% filter(Ethnicity == "African.American" & LCT == "GG")
'%nin%'<-Negate('%in%')
ffq_filt<- ffq %>% filter(SubjectID %nin% af_gg$SubjectID) %>% filter(SubjectID != 6083 & SubjectID != 8015)
plyr::count(ffq_filt, c("Ethnicity", "LCT"))
ffq <- ffq_filt
```

```{r}
summary(ffq)
```

```{r}
ffq$sex<- as.factor(ffq$sex)
ffq$Ethnicity<-as.factor(ffq$Ethnicity)
ffq$LCT<-as.factor(ffq$LCT)
ffq$LNP<-as.factor(ffq$LNP)
```

Does the frequency of LP vs LNP significantly differ among ethnicities? 

```{r}
plyr::count(ffq, c("Ethnicity", "LNP")) %>% tidyr::spread(., LNP, freq) %>% htmlTable::htmlTable(rnames = F)
```

```{r}
cnts<- plyr::count(ffq, c("Ethnicity", "LNP"))
cnts_wide <- tidyr::spread(cnts, LNP, freq)
cnts_wide[is.na(cnts_wide)]<- 0
rownames(cnts_wide)<- cnts_wide$Ethnicity
cnts_wide<- cnts_wide[-1,-1] #get rid of Ethnicity and African Americans since that is artificailly made to have LNP=0
fisher.test(as.matrix(cnts_wide)) # 2.2e-16

pairwiseNominalIndependence(as.matrix(cnts_wide), fisher = T, gtest=F, chisq=F, method = "fdr") %>% htmlTable::htmlTable(rnames = F)

```


correlation matrix
```{r}
ffqtmp<- ffq %>% dplyr::select(SubjectID, age, ht_cm:bmi_final, d_total:dt_kcal)
ffqcorr<- cor(ffqtmp[,-1])
ffqp<- ggcorrplot::cor_pmat(ffqtmp[,-1])
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/corrplot_ffq.pdf", height = 5, width = 6)
corrplot::corrplot(ffqcorr, is.corr=F, tl.srt=65, p.mat=ffqp, sig.level = 0.05, insig="pch",
                   mar=c(0,0,1,0))
dev.off()
```
```{r}
ffqcorr
ffqp
```


Summary of intake by genotype
```{r}
aa<- ffq %>% filter(LCT == "AA") %>% summary(.)
ag<- ffq %>% filter(LCT == "AG") %>% summary(.)
gg<- ffq %>% filter(LCT == "GG") %>% summary(.)

summaryBy(lacs ~ LCT, data=ffq, FUN = list(mean, max, min, median)) %>%
  htmlTable::htmlTable()
```


plots
```{r, eval =F}

ggduo(ffq, 
      mapping=aes(color=LNP),
      columnsX = c("lacs","d_total", "dt_kcal"),
      columnsY=c("LNP", "Ethnicity", "age", "sex"),
      types = list(continuous= "points", comboHorizontal="box"))

ggduo(ffq, 
      mapping=aes(color=LNP),
      columnsX = c("lacs","d_total", "dt_kcal"),
      columnsY=c("ht_cm", "weightv2", "bmi_final"),
      types = list(continuous= "points", comboHorizontal="box")) +
  theme_bw()

```


```{r plot-lactosevdtotal}
p<-ggplot(data=ffq, aes(x= d_total, y = lacs)) + geom_point(aes(shape=sex, color = LNP), size = 2) + theme_bw() + 
  ggtitle("Average Lactose Consumption vs Dairy Intake ") + ylab("Estimated Lactose Consumption (g)") +
  xlab("Total Dairy Consumption (servings)") + geom_smooth(method="lm", se=F, col= "darkgray", linetype="dashed") + 
  theme(legend.position = "bottom", 
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank())
p
ggsave(p,file = "/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/scatter_ffq_lactose_dtotal.pdf", height=5, width=6, units="in", dpi=300)
```

```{r}
a<- ggplot(data = ffq, aes(x = dt_kcal, y = lacs, color = LNP)) + geom_point(aes(shape = sex), size = 2) + 
  geom_rug() + theme_bw() + theme(legend.position = "bottom") + ggtitle("Latose (g) vs dt_kcal")
b<- ggplot(data = ffq, aes(x = dt_kcal, y = d_total, color = LNP)) + geom_point(aes(shape = sex), size = 2) + geom_rug() +theme_bw() + theme(legend.position = "bottom") + ggtitle("Total Dairy (cup eq.) vs dt_kcal")
#pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/rugplot_ffq.pdf", width = 7, height = 5)
plot_grid(a, b)
dev.off()
```
# Check out distributions

if we need to transform any variables to make model fit better, then we can just use log in the model (so we can back transform easier)
```{r}
hist(log(ffq$lacs))
hist(log(ffq$d_total))
hist(log(ffq$dt_kcal))
```

# Are there differences among sex and ethnicity? 

```{r}
make_boxplots<- function(x, y, dat){
  varx = dat[,x]
  vary = dat[,y]
  y_trans = log(dat[,y])
  p<- ggplot(data = dat, aes(x = varx, y = vary)) + geom_boxplot(aes(fill = varx), varwidth = T, notch = T) + theme_bw() +
    xlab(x) + ylab(y)
  print(p)
  fit<-lm(y_trans~varx, data = dat)
  plot(fit)
  #summary(fit) %>% print()
  car::Anova(fit) %>% print()
}
```

```{r}
##total intake does not significantly differ by ethnicities
make_boxplots(dat = ffq, y = "dt_kcal", x = "Ethnicity")#0.4031
make_boxplots(dat = ffq, y = "d_total", x = "Ethnicity") #0.02087
make_boxplots(dat = ffq, y = "lacs", x = "Ethnicity")#0.2368
##since ethnicity has super unequal groups let's do a sanity check with kruskal
kruskal.test(ffq$dt_kcal ~ ffq$Ethnicity) #0.5098
kruskal.test(ffq$d_total ~ ffq$Ethnicity) #0.0432
kruskal.test(ffq$lacs ~ffq$Ethnicity) #0.1506

make_boxplots(dat = ffq, y = "dt_kcal", x = "sex") 
#men consumed significantly more kcal than women; p-value: 1.627e-10 
make_boxplots(dat = ffq, y = "d_total", x = "sex") 
#men consumed significantly more total dairy than women; p-value = 0.0004875
make_boxplots(dat = ffq, y = "lacs", x = "sex")
#difference in consumption approached significance; p-value 5.877e-05
```

```{r}
a<- ggplot(data = ffq, aes(x = Ethnicity, y = lacs)) + geom_boxplot(aes(fill=Ethnicity)) + theme_bw() + theme(legend.position = "none", axis.text.x = element_text(angle = 45, vjust=1, hjust = 1))
b<- ggplot(data = ffq, aes(x = Ethnicity, y = d_total)) + geom_boxplot(aes(fill=Ethnicity)) + theme_bw() + theme(legend.position  = "none", axis.text.x = element_text(angle = 45, vjust=1, hjust = 1))
#pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/boxplot_ffq_lactose_dtotal_eth.pdf", width = 7, height = 5)
plot_grid(a,b)
dev.off()
```

# Does Lactose consumption significantly differ between LP and LNP (all subjects)? 
```{r}
t.test(log(lacs) ~ LNP, data = ffq)# not significantly different p = 0.1298

pdf("../plots/boxplot_ffq_lactose_lnp.pdf", width = 4, height = 4)
ggplot(data = ffq, aes(x = LNP, y = lacs)) + geom_boxplot(aes(fill = LNP), varwidth = T, notch= T) + theme_bw()
dev.off()
#ggplot(data = ffq, aes(x = LCT, y = lacs)) + geom_boxplot(aes(fill = LNP), varwidth = T, notch= T) + theme_bw()
```

is there an association when controling for kcal and ethnicity?
```{r}
lac_fit<- lm(log(lacs) ~ log(dt_kcal)+ Ethnicity +LNP, data = ffq) #we can't be sure about any LNP*Ethnicity interaction because we have taken out all GG African Americans, so we keep everything as + 
plot(lac_fit) #looks good
hist(lac_fit$residuals) #looks good
summary(lac_fit)
Anova(lac_fit)
```
```{r}
emmeans(lac_fit, pairwise ~ LNP, type= "response")
```

Does total dairy consumption significantly differ between LP and LNP subjects?

```{r}
#simple t test
t.test(log(d_total) ~ LNP, data = ffq)#0.1186

```

```{r}
dairy_fit <- lm(log(d_total) ~ log(dt_kcal) + Ethnicity +LNP, data = ffq)
plot(dairy_fit)
hist(dairy_fit$residuals)
summary(dairy_fit)
Anova(dairy_fit)
```

```{r eval = F}
a<- ggplot(data = ffq, aes(x = LNP, y = lacs)) + geom_boxplot(aes(fill = LNP), varwidth=T, notch=T ) + theme_bw() + theme(legend.position ="none")
b<- ggplot(data = ffq, aes(x = LNP, y = d_total)) + geom_boxplot(aes(fill = LNP), varwidth=T, notch=T ) + theme_bw() + theme(legend.position ="none")
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/boxplots_ffq_lac_dairy_lnp.pdf", width = 7, height = 5)
plot_grid(a, b)
dev.off()
```
Average consumption (raw data, not the estimated/adjusted mean)
```{r}
asa %>% group_by(Ethnicity, LNP) %>% 
  summarize(mean_lactose = round(mean(Lactose.consumed),2),
            sd_lactose = round(sd(Lactose.consumed),2),
            mean_dairy = round(mean(D_TOTAL),2),
            sd_dairy = round(sd(D_TOTAL),2)) %>% htmlTable::htmlTable(rnames = F)
```

```{r}
asa %>% group_by( LNP) %>% 
  summarize(mean_lactose = round(mean(Lactose.consumed),2),
            sd_lactose = round(sd(Lactose.consumed),2),
            mean_dairy = round(mean(D_TOTAL),2),
            sd_dairy = round(sd(D_TOTAL),2)) %>% htmlTable::htmlTable(rnames = F)
```

