---
title: "06_ffq_lct_analysis_caucasians_hispanics"
---

Author: E.Chin
Purpose: FFQ Lactose and d_total analysis w.r.t. LNP genotypes; for use with 16s analysis paper. This is only for the Caucasian and Hispanic participants. 


```{r setup}
library(dplyr)
library(ggplot2)
library(car)
library(doBy)
library(GGally)
library(corrplot)
library(cowplot)
library(emmeans)

ffq<- read.csv("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/out_ffq/02_ffq_lactose_dtotal_lct.csv", header=T) #these are the subjects who have lactose (average over 2 or 3 recalls) estimates available from usable ffq24 recalls  

#include only cacuasian and hispanic subjets
ffq <- ffq %>% filter(Ethnicity == "Caucasian" | Ethnicity == "Hispanic") #nrow = 260
```

```{r}
summary(ffq)
```

```{r}
ffq$sex<- as.factor(ffq$sex)
ffq$Ethnicity<-as.factor(ffq$Ethnicity)
ffq$LCT<-as.factor(ffq$LCT)
ffq$LNP<-as.factor(ffq$LNP)
```

```{r}
male<- ffq %>% filter(sex == 1)
female<- ffq %>% filter(sex == 2)
```

```{r}
summary(male)
summary(female)
```


```{r}
plyr::count(ffq, c("sex", "LNP")) %>% htmlTable::htmlTable(rnames = F)
plyr::count(ffq, c("sex", "LCT")) %>% htmlTable::htmlTable(rnames=F)
```



correlation matrix

```{r}
makecorrs<- function(data, titlename){
  asatmp<- data %>% dplyr::select(SubjectID, age, ht_cm:bmi_final, d_total:dt_kcal)
  asacorr<- cor(asatmp[,-1])
  title = paste0("corrplot for ", titlename)
  asap<- ggcorrplot::cor_pmat(asatmp[,-1])
  cplot<- corrplot::corrplot(asacorr, is.corr=F, tl.srt=65, p.mat=asap, sig.level = 0.05, insig="pch",
                   mar=c(0,0,3,0), title = title)
  return(cplot)
}
```

```{r}
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/corrplot_ffq_men.pdf", width = 4, height =4)
corr_men<- makecorrs(male, "men")
dev.off()

pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/corrplot_ffq_women.pdf", width = 4, height =4)
corr_women<- makecorrs(female, "women")
dev.off()
```
```{r}
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/scatter_ffq_lactose_dtotal_sex.pdf", width = 6, height =4)
ggplot(data = ffq, aes(x = d_total, y = lacs, color = LNP)) + geom_point(aes(fill =LNP), size = 2) + 
  geom_rug()+ geom_smooth(method = "lm", se = F) + 
  facet_wrap(~factor(sex), scales = "free") + theme_bw()
dev.off()
```

# Check out distributions

```{r}
hist(male$lacs %>% log())
hist(male$d_total %>% log())
hist(female$lacs %>% log()) #still a little skewed but we can see how model residuals look 
hist(female$d_total %>% log())#still a little skewed but we can see how model residuals look 
```
```{r}
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/boxplot_ffq_lactose_lct_sex.pdf", width = 6, height =4)
ggplot(data = ffq, aes(x= LCT, y = lacs)) + geom_boxplot(aes(fill = LNP), varwidth = T, notch = T) + ggtitle("Lactose")+
  theme_bw() + facet_wrap(~sex)
dev.off()
```

```{r}
pdf("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/plots/boxplot_ffq_dairy_lct_sex.pdf", width = 6, height =4)
ggplot(data = ffq, aes(x= LCT, y = d_total)) + geom_boxplot(aes(fill = LNP), varwidth = T, notch = T) + 
  theme_bw() + facet_wrap(~sex) + ggtitle("Total Dairy")
dev.off()
```



# Does Lactose consumption significantly differ between LP and LNP? 
```{r message= F}
ix_plots<- function(data, varx, vary){
  plot_x<- data[,varx]
  plot_y <- data[,vary]
  q<- ggplot(data = data, aes(x = plot_x, y = plot_y, color = LNP)) + geom_point(aes(fill = LNP)) + 
    geom_smooth(method = "lm", se = F, aes(color = LNP)) + theme_bw() + xlab(varx) + ylab(vary)
  print(q)
}

ix_plots(data = male, varx = "dt_kcal", vary="lacs")
ix_plots(data = male, varx = "dt_kcal", vary="d_total")
ix_plots(data = female, varx = "dt_kcal", vary="lacs")
ix_plots(data = female, varx = "dt_kcal", vary="d_total")
```

Does D_TOTAL or Lactose consumption differ with LNP status in MEN?
```{r}
t.test(data = male, log(d_total) ~ LNP, alternative = "two.sided") #0.5463
t.test(data = male, log(lacs) ~ LNP, alternative = "two.sided") #0.2176
```

```{r}
do_lm<- function(df, yvar, x_main, x_cov){
  y = df[,yvar] %>% log()
  xvar = df[,x_main]
  xcov= df[,x_cov] %>% log()
  fit_ix<- lm(y ~ xcov*xvar, data = df)
  s<- (summary(fit_ix))
  d<- s$coefficients %>% as.data.frame()
  if(d$`Pr(>|t|)`[4] < 0.05){
    plot(fit_ix)
    hist(fit_ix$residuals)
    print(s)
  }else if(d$`Pr(>|t|)`[4] >= 0.05){
    print("no significant interction")
    fit<- lm(y ~ xcov + xvar, data = df)
    plot(fit)
    hist(fit$residuals)
    car::Anova(fit) %>% print()
  }

}
```

```{r}
do_lm(df = male, yvar = "lacs", x_main = "LNP", x_cov = "dt_kcal") #0.1695
```
```{r}
do_lm(df = male, yvar = "d_total", x_main = "LNP", x_cov = "dt_kcal") #0.4373

```

```{r}
do_lm(df = female, yvar = "lacs", x_main = "LNP", x_cov = "dt_kcal") #0.2233

```

```{r}
do_lm(df = female, yvar = "d_total", x_main = "LNP", x_cov = "dt_kcal")#0.1495
```
Are there any differences w.r.t. LCT genoytpe? 

```{r}
car::Anova(lm(log(d_total) ~ LCT, data = male))#0.6871
car::Anova(lm(log(lacs) ~ LCT, data = male))#0.229

do_lm(df = male, yvar = "d_total", x_main = "LCT", x_cov = "dt_kcal") #0.7298
do_lm(df = male, yvar = "lacs", x_main = "LCT", x_cov = "dt_kcal") #0.2337
```

```{r}
car::Anova(lm(log(d_total) ~ LCT, data = female))#0.2652
car::Anova(lm(log(lacs) ~ LCT, data = female))#0.2174

do_lm(df = female, yvar = "d_total", x_main = "LCT", x_cov = "dt_kcal") #0.123
do_lm(df = female, yvar = "lacs", x_main = "LCT", x_cov = "dt_kcal") #0.131s
```


