---
title: "05_get_lactose"
output: 
---

1) remove SUBJECTS who report protein powder, meal replacement powders, or cocoa powder/hot cocoa in ANY recall because the amount of lactose can vary wildly depending on the specific entry 

2) export D_TOTAL (raw ASA24 variable, incl calcium foritifed soy milk) with the lactose estimates 

refer to excluded foodcode spreadsheets for details

EC 02/04/2022 note that this is an updated version of /FL100_ASA24_lactose_estimation/06* and uses data downloaded from redcap 02/04/2022, and **starts** with all subjects who were enrolled (not only those who finished v2); results are the same though but this is 'more correct' for making a flow chart of subjects inclusion/exclusion


```{r}
library(dplyr)
setwd("/Users/elizabeth.chin/Desktop/milk/FL100_ASA24_lactose_estimation/")
#get the items
items<- read.csv("data/raw_inputs/Items_JACorrections.csv", header=T) #22391
inf<- read.csv("data/raw_inputs/INFMYPHEI_JA_Corrections.csv", header=T) #5387
items$UserName<- gsub("[^0-9.-]", "", items$UserName)
inf$UserName<- gsub("[^0-9.-]", "", inf$UserName)

#import the lactose estimations
lac16<- read.csv("data/outputs/final_matches/final_matches16.csv",header=T) #2282 
lac14<- read.csv("data/outputs/final_matches/final_matches14.csv",header=T) #1167

#bring in the recall totals (so we can kcal and qc filter)
#these are for everyone who renrolled (n = 393)
r2<- read.csv("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/raw_asa24/recall2_CTSC24532USDAWHNRCNu_DATA_2022-02-04_1347.csv", header=T)
r3<- read.csv("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/raw_asa24/recall3_CTSC24532USDAWHNRCNu_DATA_2022-02-04_1349.csv", header=T)
r4<- read.csv("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/raw_asa24/recall4_CTSC24532USDAWHNRCNu_DATA_2022-02-04_1350.csv", header=T)

#these are for only those who completed v2
#r1<- read.csv("data/raw_inputs/CTSC24532USDAWHNRCNu-ASA24T_DATA_2020-01-02_1127.csv", header=T)
#r2<- read.csv("data/raw_inputs/CTSC24532USDAWHNRCNu-ASA241_DATA_2020-01-02_1128.csv", header=T)
#r3<- read.csv("data/raw_inputs/CTSC24532USDAWHNRCNu-ASA242_DATA_2020-01-02_1128.csv", header=T)
#r4<- read.csv("data/raw_inputs/CTSC24532USDAWHNRCNu-ASA243_DATA_2020-01-02_1129.csv", header=T)

#LCT
lct<- read.csv("/Users/elizabeth.chin/Desktop/milk/genotype_ethnicity/FL100_FINAL_LCT_genotype_clean.csv", header=T)

#metadata 
eth<- read.csv("/Users/elizabeth.chin/Desktop/milk/genotype_ethnicity/FL100_FINAL_ethnicities_clean.csv", header=T)
bins<-read.csv("/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/CTSC24532USDAWHNRCNu-BinAgeSexBMIallSubje_DATA_2022-02-09_1449.csv", header=T)

#protein powder food codes to flag for exclusion
ex14<- read.csv("data/raw_inputs/exclude_foodcode14.csv", header=T)
ex16<- read.csv("data/raw_inputs/exclude_foodcode16.csv", header=T)
```
# Merge items with the lactose estimations

2016:  
```{r}
items$code<- paste0(items$FoodCode, "_", items$Food_Description) %>% trimws()
lac16$code<- paste0(lac16$FoodCode, "_", lac16$Food_Description) %>% trimws()
items.lac<- merge(items, lac16, by = "code", all.x = T)
sum(is.na(items.lac$Lactose))
#View(items.lac %>% filter(is.na(Lactose) & FoodComp != 2))
#View(items.lac %>% filter(is.na(Lactose)))

#what doesn't have a food amount associated with it? 
sum(is.na(items.lac$FoodAmt))
noamt<- items.lac %>% filter(is.na(FoodAmt)) #221 foods but many are FoodCode == 9
noamt<- items.lac %>% filter(is.na(FoodAmt) & FoodCode.x != 9)#only 16 foods- these are foods where subjects entered the item but no further details so nutrients could not be calculated
#find out if any of these foods are high lactose containing? 
df<- data.frame(items$FoodCode, items$Food_Description)
#View(df %>% filter(items.FoodCode %in% noamt$FoodCode.x) %>% unique()) #only food with lactose is 92161005 	Cappuccino, sweetened Subj 8029 R2, but this recall is not contained in r2 (I suppose this subject didn't finish V2?)
```

2014
```{r}
inf$code<- paste0(inf$FoodCode, "_", inf$Food_Description) %>% trimws()
lac14$code<- paste0(lac14$FoodCode, "_", lac14$Food_Description) %>% trimws()
inf.lac<- merge(inf, lac14, by = "code", all.x=T)
sum(is.na(inf.lac$Lactose)) #31
sum(is.na(inf.lac$FoodAmt)) #54
noamt<- inf.lac %>% filter(is.na(FoodAmt)) %>% select(NCC.Food.Description, everything())
df<- data.frame(inf$FoodCode, inf$Food_Description) #the reason why there are some foods in lac14 without a food description (but not in lac16) is bceause when I was retrieving the ASA24 unique food list, I did na.omit for 2016 users but not 2014. 

```

make your life easier, rbind 2014 and 2016
```{r}
inf.lac2<- inf.lac %>% select(UserName, RecallNo, FoodCode.x, Food_Description.x, FoodAmt, Lactose, NCC.Food.Description, D_TOTAL,KCAL)
items.lac2<- items.lac %>% select(UserName, RecallNo, FoodCode.x, Food_Description.x, FoodAmt, Lactose, NCC.Food.Description, D_TOTAL, KCAL)

all.lac<- rbind(inf.lac2, items.lac2)
colnames(all.lac)<-c("SubjectID", "RecallNo", "FoodCode", "Food_Description", "FoodAmt", "Lactose.p100g", "NCC.Food.Description", "D_TOTAL", "KCAL")

#convert lactose to per-g basis
all.lac$Lactose.p1g<- all.lac$Lactose.p100g/100
```

# Get lactose amounts
```{r}
all.lac$Lactose.consumed<- all.lac$Lactose.p1g * all.lac$FoodAmt
```

#who is reporting protein powders/drinks

```{r}
pro.codes<-rbind(ex14, ex16)
pro<- all.lac %>% filter(FoodCode %in% pro.codes$FoodCode)

dim(unique(pro[,1:2])) #116 unique subjects x recalls reporting protein powders/drinks (75 are from at-home recalls)
length(unique(pro[,1])) #59 subjects
```

# Get only usable recalls

```{r}
'%nin%'<-Negate('%in%')
#remove the training recall & subjects'recalls reporting protein
all.nt<- all.lac %>% filter(RecallNo != 1) #20846 rows total, 1121 unique subjectxrecalls
r2$sr.code<- paste0(r2$subject_id, "_", r2$recallno_tnf2)
r3$sr.code<- paste0(r3$subject_id, "_", r3$recallno_tnf3)
r4$sr.code<- paste0(r4$subject_id, "_", r4$recallno_tnf4)

r2.ok<- r2 %>% filter(recallqc_passed_2 == 1) %>% 
  select(subject_id, recallno_tnf2) #384
r3.ok<- r3 %>% filter(recallqc_passed_3 == 1) %>% 
  select(subject_id, recallno_tnf3) #378
r4.ok<- r4 %>% filter(recallqc_passed_4 == 1) %>% 
  select(subject_id, recallno_tnf4) #346
colnames(r2.ok)<-c("SubjectID", "RecallNo")
colnames(r3.ok)<-colnames(r2.ok)
colnames(r4.ok)<-colnames(r2.ok)
recalls<-rbind(r2.ok, r3.ok, r4.ok)

who<- plyr::count(recalls, "SubjectID")
keep<- who %>% filter(freq > 1) #these are the subjects with at least 2 at-home recalls passing ASA24 QC

r2.keep<- r2.ok %>% filter(SubjectID %in% keep$SubjectID) #377
r3.keep<- r3.ok %>% filter(SubjectID %in% keep$SubjectID) #378
r4.keep<- r4.ok %>% filter(SubjectID %in% keep$SubjectID) #346

check<- list(r2= r2.keep$SubjectID,
             r3= r3.keep$SubjectID, 
             r4= r4.keep$SubjectID)
gplots::venn(check) 

check2<- list(unique(pro$SubjectID), r2.ok$SubjectID, r3.ok$SubjectID, r4.ok$SubjectID)
gplots::venn(check2)
```
# retain subjects items for only QC'd recalls  
```{r}
r2.lac<- all.nt %>% filter(RecallNo == 2) %>% filter(SubjectID %in% r2.keep$SubjectID)
length(unique(r2.lac$SubjectID))#377
r3.lac<- all.nt %>% filter(RecallNo == 3) %>% filter(SubjectID %in% r3.keep$SubjectID)
length(unique(r3.lac$SubjectID)) #378
r4.lac<- all.nt %>% filter(RecallNo == 4) %>% filter(SubjectID %in% r4.keep$SubjectID)
length(unique(r4.lac$SubjectID)) #346

qc.lac<- rbind(r2.lac, r3.lac, r4.lac)
dim(qc.lac)
length(unique(qc.lac$SubjectID)) #378 subjects
colnames(qc.lac)
```

# Calculate the *total* Lactose and D_TOTAL *per recall per subject*

```{r}
per.recall<- aggregate(cbind(Lactose.consumed, D_TOTAL, KCAL) ~ SubjectID + RecallNo, data=qc.lac, FUN="sum")
dim(per.recall) #1101
#check
dim(per.recall %>% select(SubjectID, RecallNo) %>% unique())
sum(keep$freq) #checks out 
```


# Calculate the *average* Lactose and D_TOTAL *per subject, averaged over n recalls*

```{r}
per.subject<- aggregate(cbind(Lactose.consumed,D_TOTAL) ~ SubjectID, data=per.recall, FUN="mean") #296 subjects
write.csv(per.subject, "/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/out_asa24/01_asa24_lactose_dtotal.csv", row.names=F)
write.csv(per.recall, "/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/out_asa24/01b_asa24_lactose_dtotal_per_recall.csv", row.names=F)

```

# Get rid of subjects reporting protein powders or cocoa powders

how many subjects would there be if we removed those who report hot cocoa? 
11830160
11830165
11830110


```{r}
'%nin%'<-Negate('%in%')
noprotein<- per.subject %>% filter(SubjectID %nin% pro$SubjectID)
dim(noprotein) #321
```

```{r}
cocoa<- all.lac %>% filter(FoodCode == 11830165 | FoodCode == 11830160 | FoodCode == 11830110)
nococoa<- noprotein %>% filter(SubjectID %nin% cocoa$SubjectID) #317 subjects 

write.csv(nococoa, "/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/out_asa24/02_asa24_lactose_dtotal_no_powders.csv", row.names=F)

#gplots::venn(list("cocoa" = unique(cocoa$SubjectID), "pro"= unique(pro$SubjectID), "nofilt" = unique(per.subject$SubjectID)))
```


# merge with sequencing, lct, ethnicity data
```{r}
master<- merge(merge(lct, eth, by = "SubjectID"), bins, by.x = "SubjectID", by.y="subject_id")%>% 
  dplyr::select(-c(visit1_date:age_ffq))#349 subjects

#lactose with metadata, protein/breakfast/cocoa powder consumers removed
master.lac<- merge(master, nococoa, by = "SubjectID") #286 subjects
write.csv(master.lac, "/Users/elizabeth.chin/Desktop/milk/FL100_lactose_lct_16s/data/03_asa24_lactose_dtotal_no_powders_lct.csv", row.names=F)

gplots::venn(list("meta" = unique(master$SubjectID), "lac" = unique(nococoa$SubjectID)))
```


```{r}
summary(master.lac)
for.plot<- master.lac
library(plotly)
shapes<- c("circle", "cross")
for.plot$shapes<- shapes[for.plot$sex]
for.plot$col<- rainbow(3)[for.plot$LCT]

plot_ly(data=master.lac, x = ~Lactose.consumed, type="scatter",
        text=~paste("Subject:",SubjectID,
                    "<br>Ethnicity:", Ethnicity,
                    "<br>Age:", age,
                    "<br>BMI:", bmi_final),
        color=~LCT, 
        symbol= ~sex,
        size=3)

```
```{r}
hist(master.lac$Lactose.consumed)
ggplot(data=master.lac, aes(x = LCT, y = Lactose.consumed)) + geom_boxplot(varwidth=T, notch=T, aes(fill=LCT)) + 
  theme_classic() + ggtitle("Lactose (g, averaged over 2 or 3 recalls) for the LCT genotype")
ggplot(data=master.lac, aes(x = as.factor(sex), y = Lactose.consumed)) + geom_boxplot(varwidth=T, notch=T, aes(fill=sex)) + 
  theme_classic() + ggtitle("Lactose (g, averaged over 2 or 3 recalls) for\nmen and women")
```

```{r}
## Summarizes data.
## Gives count, mean, standard deviation, standard error of the mean, and confidence 
## interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE, conf.interval=.95) {
    library(doBy)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # Collapse the data
    formula <- as.formula(paste(measurevar, paste(groupvars, collapse=" + "), sep=" ~ "))
    datac <- summaryBy(formula, data=data, FUN=c(length2,mean,sd), na.rm=na.rm)

    # Rename columns
    names(datac)[ names(datac) == paste(measurevar, ".mean",    sep="") ] <- measurevar
    names(datac)[ names(datac) == paste(measurevar, ".sd",      sep="") ] <- "sd"
    names(datac)[ names(datac) == paste(measurevar, ".length2", sep="") ] <- "N"
    
    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
    
    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult
    
    return(datac)
}
```

```{r}
df<- summarySE(master.lac, measurevar="Lactose.consumed", 
                  groupvars="LCT")

ggplot(df, aes(x= LCT, y = Lactose.consumed)) + geom_bar(aes(fill=LCT), stat="identity") +
  geom_errorbar(aes(ymin=Lactose.consumed-se, ymax=Lactose.consumed + se), width=0.2) +
  theme_classic() + ggtitle("Lactose consumed (g, averaged over 2 or 3 recalls)\nby LCT Genotype, excluding protein powder consumers") +
  xlab("LCT Genotype") + ylab("Lactose Consumed (g +/- SEM)")
#ggsave("/Users/elizabeth.chin/Desktop/milk/FL100_ASA24_lactose_estimation/plots/lactose_by_LCT.png", width=6, height=5)
```


